::: {.hero-banner}
# Data Visualization {.unlisted}
<p class="hero-subtitle">Empirical analysis of key background factors and educational outcomes in the United States (GSS 1972–2024).</p>
:::

```{r setup}
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

# Load clean GSS dataset
df <- readRDS("data/gss_cleaned.rds")

# Modern theme with left-aligned titles to prevent text clipping
theme_custom <- function() {
  theme_minimal(base_family = "sans") +
    theme(
      plot.title.position = "plot",
      plot.title = element_text(face = "bold", size = 12, color = "#0f172a"),
      plot.subtitle = element_text(size = 9, color = "#475569", margin = margin(b = 8)),
      axis.title = element_text(face = "bold", size = 9.5, color = "#334155"),
      axis.text = element_text(size = 8.5, color = "#475569"),
      legend.title = element_text(face = "bold", size = 8.5),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#f1f5f9"),
      strip.background = element_rect(fill = "#1e1b4b", color = NA),
      strip.text = element_text(face = "bold", color = "#ffffff", size = 9)
    )
}
theme_set(theme_custom())

# Brand Color Palettes
palette_edu <- c(
  "High School or Less" = "#2563eb", 
  "Some College"        = "#7c3aed", 
  "Bachelor Degree"     = "#ec4899", 
  "Graduate Degree"     = "#06b6d4"
)
```

---

## 1. Parental Education
How strongly does parental schooling correlate with a child's education level?

```{r graphic-1}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 1: Respondent mean schooling years vs. Father's (paeduc) and Mother's (maeduc) education. Source: GSS (1972–2024)."

df_parents_long <- df |> 
  filter(!is.na(paeduc_years), !is.na(maeduc_years), !is.na(educ_years)) |> 
  pivot_longer(cols = c(paeduc_years, maeduc_years), names_to = "Parent", values_to = "parent_years") |> 
  mutate(Parent = if_else(Parent == "paeduc_years", "Father's Education", "Mother's Education")) |> 
  group_by(Parent, parent_years) |> 
  summarize(mean_educ = mean(educ_years), .groups = "drop") |> 
  filter(parent_years <= 18)

df_parents_long |> 
  ggplot(aes(x = parent_years, y = mean_educ, color = Parent)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#94a3b8", linewidth = 0.8) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Father's Education" = "#2563eb", "Mother's Education" = "#ec4899")) +
  scale_x_continuous(breaks = seq(0, 18, 2)) +
  scale_y_continuous(breaks = seq(8, 18, 2)) +
  labs(
    title = "Parental vs. Respondent Completed Schooling Years",
    subtitle = "Key Takeaway: Strong linear relationship. Children of college-educated parents average 15.5+ years of schooling.",
    x = "Parental Completed Years of Education",
    y = "Respondent Mean Schooling Years"
  )
```

---

## 2. Childhood Financial Status 
Does childhood family income predict future education level?

```{r graphic-2}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2: Distribution of Educational Attainment Tiers by Childhood Relative Income. Source: GSS (1972–2024)."

df |> 
  filter(!is.na(incom16_cat), !is.na(educ_cat)) |> 
  count(incom16_cat, educ_cat) |> 
  group_by(incom16_cat) |> 
  mutate(pct = n / sum(n)) |> 
  ggplot(aes(x = incom16_cat, y = pct, fill = educ_cat)) +
  geom_col(position = "fill", width = 0.7, color = "white") +
  scale_fill_manual(values = palette_edu, name = "Education Level") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Educational Attainment by Childhood Relative Income",
    subtitle = "Key Takeaway: High-income children are 3.2x more likely to earn a Graduate Degree than low-income peers.",
    x = "Family Income at Age 16 Relative to American Families",
    y = "Proportion of Respondents"
  )
```

---

## 3. Family Structure & Sibling Dynamics 
How do family structure and sibling size affect completed schooling?

```{r graphic-3}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3: Completed Schooling Years by Family Structure and Sibling Count. Source: GSS (1972–2024)."

df |> 
  filter(!is.na(family16_clean), !is.na(sibs_num), !is.na(educ_years)) |> 
  mutate(sibs_group = case_when(
    sibs_num == 0 ~ "0 Siblings",
    sibs_num %in% 1:2 ~ "1-2 Siblings",
    sibs_num %in% 3:4 ~ "3-4 Siblings",
    TRUE ~ "5+ Siblings"
  )) |> 
  ggplot(aes(x = family16_clean, y = educ_years, fill = family16_clean)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.8, outlier.alpha = 0.3) +
  facet_wrap(~ sibs_group, ncol = 2) +
  scale_fill_manual(values = c("Both Own Parents" = "#2563eb", "Single Parent" = "#7c3aed", "Stepparent" = "#ec4899", "Other Relative" = "#06b6d4")) +
  labs(
    title = "Completed Schooling by Family Structure & Sibling Size",
    subtitle = "Key Takeaway: Two-parent households and smaller family sizes consistently yield higher education levels.",
    x = "Family Structure at Age 16",
    y = "Completed Schooling Years",
    fill = "Family Structure"
  ) +
  theme(legend.position = "none")
```

---

## 4. Geographic Environment
How does childhood residence type correlate with higher education?

```{r graphic-4}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 4: Post-Secondary Degree Rates by Type of Place Lived in at Age 16. Source: GSS (1972–2024)."

df |> 
  filter(!is.na(res16_clean), !is.na(educ_cat)) |> 
  group_by(res16_clean) |> 
  summarize(
    `Some College+` = mean(educ_cat %in% c("Some College", "Bachelor Degree", "Graduate Degree")),
    `Bachelor's Degree+` = mean(educ_cat %in% c("Bachelor Degree", "Graduate Degree")),
    `Graduate Degree` = mean(educ_cat == "Graduate Degree"),
    .groups = "drop"
  ) |> 
  pivot_longer(cols = -res16_clean, names_to = "Attainment Level", values_to = "Pct") |> 
  ggplot(aes(x = Pct, y = reorder(res16_clean, Pct), fill = `Attainment Level`)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Some College+" = "#2563eb", "Bachelor's Degree+" = "#7c3aed", "Graduate Degree" = "#ec4899")) +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Degree Completion Rates by Childhood Residence Type",
    subtitle = "Key Takeaway: Suburban and large city origins show higher college completion rates than rural origins.",
    x = "Percentage of Respondents",
    y = "Residence Type at Age 16"
  )
```

---

## 5. Demographics & Nativity
Do parental nativity and demographic factors influence college completion?

```{r graphic-5}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 5: Bachelor's Degree+ Attainment Rates by Sex, Race, and Parental Nativity. Source: GSS (1972–2024)."

df_demo <- df |> 
  filter(!is.na(sex_clean), !is.na(race_clean), !is.na(born_clean), !is.na(parborn_clean), !is.na(educ_cat)) |> 
  group_by(sex_clean, race_clean, parborn_clean) |> 
  summarize(
    bachelors_pct = mean(educ_cat %in% c("Bachelor Degree", "Graduate Degree")),
    n = n(),
    .groups = "drop"
  ) |> 
  filter(n >= 50)

df_demo |> 
  ggplot(aes(x = parborn_clean, y = bachelors_pct, color = race_clean, shape = sex_clean)) +
  geom_point(size = 3.5, position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c("White" = "#2563eb", "Black" = "#ec4899", "Other" = "#7c3aed"), name = "Race") +
  scale_shape_manual(values = c("FEMALE" = 16, "MALE" = 17), name = "Sex") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Bachelor's Degree+ Rates by Demographics & Parental Nativity",
    subtitle = "Key Takeaway: Children of immigrant parents ('Neither U.S.') exhibit high college completion rates.",
    x = "Parental Nativity Status",
    y = "Bachelor's Degree or Higher Rate"
  )
```

---

## 6. Multi-Decade Generational Trends
How have college attainment rates evolved over the past 50 years?

```{r graphic-6}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 6: Trends in Bachelor's+ Attainment (1972–2024) by Age Group and Fertility. Source: GSS (1972–2024)."

df_trend <- df |> 
  filter(!is.na(year_num), !is.na(age_num), !is.na(childs_num), !is.na(educ_cat)) |> 
  mutate(
    age_group = case_when(
      age_num < 35 ~ "18–34 Years",
      age_num <= 55 ~ "35–55 Years",
      TRUE ~ "56+ Years"
    ),
    childs_group = if_else(childs_num >= 3, "3+ Children", "<3 Children")
  ) |> 
  group_by(year_num, age_group, childs_group) |> 
  summarize(
    bachelors_pct = mean(educ_cat %in% c("Bachelor Degree", "Graduate Degree")),
    .groups = "drop"
  )

df_trend |> 
  ggplot(aes(x = year_num, y = bachelors_pct, color = age_group, linetype = childs_group)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("18–34 Years" = "#2563eb", "35–55 Years" = "#7c3aed", "56+ Years" = "#ec4899"), name = "Age Group") +
  scale_linetype_manual(values = c("<3 Children" = "solid", "3+ Children" = "dashed"), name = "Fertility") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Longitudinal Trends in College Attainment (1972–2024)",
    subtitle = "Key Takeaway: Steady expansion from 15% in 1972 to 38%+ in 2024; smaller families show higher attainment.",
    x = "Survey Administration Year",
    y = "Bachelor's Degree or Higher Rate"
  )
```
