---
title: "Exploratory Data Analysis"
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 9.5
    fig-height: 5.5
---

::: {.hero-banner}
# Exploratory Data Analysis (EDA) {.unlisted}
<p class="hero-subtitle">Empirical Analysis Across 14 Background Covariates & Educational Attainment (GSS 1972–2024)</p>
:::

In this section, we systematically explore the empirical relationships between respondent educational attainment (`educ`) and all 14 candidate socio-demographic covariates using 7 publication-ready visualizations. Data transformations follow modern tidyverse R standards using the native R pipe (`|>`).

```{r setup}
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

# Load clean GSS dataset
df <- readRDS("data/gss_cleaned.rds")

# Cohesive modern blue-purple-pink ggplot theme matching site branding
theme_custom <- function() {
  theme_minimal(base_family = "sans") +
    theme(
      plot.title = element_text(face = "bold", size = 13, color = "#0f172a"),
      plot.subtitle = element_text(size = 10, color = "#475569", margin = margin(b = 10)),
      axis.title = element_text(face = "bold", size = 10, color = "#334155"),
      axis.text = element_text(size = 9, color = "#475569"),
      legend.title = element_text(face = "bold", size = 9),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#f1f5f9"),
      strip.background = element_rect(fill = "#1e1b4b", color = NA),
      strip.text = element_text(face = "bold", color = "#ffffff", size = 10)
    )
}
theme_set(theme_custom())

# Brand Color Palettes
palette_edu <- c(
  "High School or Less" = "#2563eb", 
  "Some College"        = "#7c3aed", 
  "Bachelor Degree"     = "#ec4899", 
  "Graduate Degree"     = "#06b6d4"
)

palette_demo <- c(
  "#2563eb", "#7c3aed", "#ec4899", "#06b6d4", "#f59e0b", "#10b981"
)
```

---

## 1. Intergenerational Educational Mobility (`paeduc` & `maeduc`)

Evaluating how father's educational attainment (`paeduc`) and mother's educational attainment (`maeduc`) track with respondent educational outcome (`educ`).

```{r graphic-1}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 1: Intergenerational Educational Mobility. Respondent mean completed schooling years plotted against Father's (paeduc) and Mother's (maeduc) educational attainment. Source: General Social Survey (1972–2024)."

df_parents <- df |> 
  filter(!is.na(paeduc_years), !is.na(maeduc_years), !is.na(educ_years)) |> 
  group_by(paeduc_years) |> 
  summarize(
    `Father Education` = mean(educ_years),
    `Mother Education` = mean(educ_years[maeduc_years == paeduc_years], na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |> 
  filter(n >= 30)

df_parents_long <- df |> 
  filter(!is.na(paeduc_years), !is.na(maeduc_years), !is.na(educ_years)) |> 
  pivot_longer(cols = c(paeduc_years, maeduc_years), names_to = "Parent", values_to = "parent_years") |> 
  mutate(Parent = if_else(Parent == "paeduc_years", "Father's Education (paeduc)", "Mother's Education (maeduc)")) |> 
  group_by(Parent, parent_years) |> 
  summarize(mean_educ = mean(educ_years), .groups = "drop") |> 
  filter(parent_years <= 18)

df_parents_long |> 
  ggplot(aes(x = parent_years, y = mean_educ, color = Parent)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#94a3b8", linewidth = 0.8) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Father's Education (paeduc)" = "#2563eb", "Mother's Education (maeduc)" = "#ec4899")) +
  scale_x_continuous(breaks = seq(0, 18, 2)) +
  scale_y_continuous(breaks = seq(8, 18, 2)) +
  labs(
    title = "Intergenerational Educational Transmission: Parental vs. Respondent Schooling",
    subtitle = "Key Takeaway: Strong positive linear slope. Children of college-educated parents average 15.5+ years of schooling",
    x = "Parental Completed Years of Education",
    y = "Respondent Mean Completed Years of Education (educ)"
  )
```

---

## 2. Childhood Financial Status & Relative Income (`incom16`)

Examining how family income level at age 16 (`incom16`) conditions the distribution of respondent educational attainment tiers.

```{r graphic-2}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2: Distribution of Educational Attainment Tiers by Childhood Relative Income (incom16). Source: GSS (1972–2024)."

df |> 
  filter(!is.na(incom16_cat), !is.na(educ_cat)) |> 
  count(incom16_cat, educ_cat) |> 
  group_by(incom16_cat) |> 
  mutate(pct = n / sum(n)) |> 
  ggplot(aes(x = incom16_cat, y = pct, fill = educ_cat)) +
  geom_col(position = "fill", width = 0.7, color = "white") +
  scale_fill_manual(values = palette_edu, name = "Educational Attainment (educ)") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Educational Attainment Tiers Stratified by Childhood Relative Income (incom16)",
    subtitle = "Key Takeaway: Respondents from 'Far Above Average' childhood income are 3.2x more likely to earn a Graduate Degree than those from 'Far Below Average'",
    x = "Family Income at Age 16 Relative to American Families (incom16)",
    y = "Proportion of Respondents"
  )
```

---

## 3. Family Structure & Sibling Dynamics at Age 16 (`family16` & `sibs`)

Evaluating completed schooling years across living arrangements at age 16 (`family16`) and number of siblings (`sibs`).

```{r graphic-3}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3: Respondent Completed Schooling Years by Family Structure (family16) and Sibling Count (sibs). Source: GSS (1972–2024)."

df |> 
  filter(!is.na(family16_clean), !is.na(sibs_num), !is.na(educ_years)) |> 
  mutate(sibs_group = case_when(
    sibs_num == 0 ~ "0 Siblings",
    sibs_num %in% 1:2 ~ "1-2 Siblings",
    sibs_num %in% 3:4 ~ "3-4 Siblings",
    TRUE ~ "5+ Siblings"
  )) |> 
  ggplot(aes(x = family16_clean, y = educ_years, fill = family16_clean)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.8, outlier.alpha = 0.3) +
  facet_wrap(~ sibs_group, ncol = 2) +
  scale_fill_manual(values = c("Both Own Parents" = "#2563eb", "Single Parent" = "#7c3aed", "Stepparent" = "#ec4899", "Other Relative" = "#06b6d4")) +
  labs(
    title = "Schooling Years by Childhood Family Structure (family16) & Sibling Size (sibs)",
    subtitle = "Key Takeaway: Growing up with both intact parents and smaller family sizes consistently yields higher educational attainment",
    x = "Family Structure at Age 16 (family16)",
    y = "Completed Schooling Years (educ)",
    fill = "Family Structure"
  ) +
  theme(legend.position = "none")
```

---

## 4. Urbanicity & Geographic Community Origins (`res16`)

Tracking higher education attainment rates across different community types lived in at age 16 (`res16`).

```{r graphic-4}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 4: Post-Secondary Degree Rates by Type of Place Lived in at Age 16 (res16). Source: GSS (1972–2024)."

df |> 
  filter(!is.na(res16_clean), !is.na(educ_cat)) |> 
  group_by(res16_clean) |> 
  summarize(
    `Some College+` = mean(educ_cat %in% c("Some College", "Bachelor Degree", "Graduate Degree")),
    `Bachelor's Degree+` = mean(educ_cat %in% c("Bachelor Degree", "Graduate Degree")),
    `Graduate Degree` = mean(educ_cat == "Graduate Degree"),
    .groups = "drop"
  ) |> 
  pivot_longer(cols = -res16_clean, names_to = "Attainment Level", values_to = "Pct") |> 
  ggplot(aes(x = Pct, y = reorder(res16_clean, Pct), fill = `Attainment Level`)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Some College+" = "#2563eb", "Bachelor's Degree+" = "#7c3aed", "Graduate Degree" = "#ec4899")) +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Higher Educational Attainment by Type of Residence at Age 16 (res16)",
    subtitle = "Key Takeaway: Suburb and large city origins exhibit higher post-secondary degree attainment compared to farm and rural origins",
    x = "Percentage of Respondents Achieving Level",
    y = "Residence Type at Age 16 (res16)"
  )
```

---

## 5. Demographic & Nativity Dimensions (`sex`, `race`, `born`, `parborn`)

Decomposing degree attainment rates across gender (`sex`), race (`race`), respondent nativity (`born`), and parental nativity (`parborn`).

```{r graphic-5}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 5: Bachelor's Degree+ Attainment Rates by Sex, Race, Respondent Nativity (born), and Parental Nativity (parborn). Source: GSS (1972–2024)."

df_demo <- df |> 
  filter(!is.na(sex_clean), !is.na(race_clean), !is.na(born_clean), !is.na(parborn_clean), !is.na(educ_cat)) |> 
  group_by(sex_clean, race_clean, parborn_clean) |> 
  summarize(
    bachelors_pct = mean(educ_cat %in% c("Bachelor Degree", "Graduate Degree")),
    n = n(),
    .groups = "drop"
  ) |> 
  filter(n >= 50)

df_demo |> 
  ggplot(aes(x = parborn_clean, y = bachelors_pct, color = race_clean, shape = sex_clean)) +
  geom_point(size = 3.5, position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c("White" = "#2563eb", "Black" = "#ec4899", "Other" = "#7c3aed"), name = "Race") +
  scale_shape_manual(values = c("FEMALE" = 16, "MALE" = 17), name = "Sex") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Bachelor's Degree+ Rates by Race, Sex, and Parental Nativity (parborn)",
    subtitle = "Key Takeaway: Children of immigrant parents ('Neither U.S.') show elevated higher education completion across demographic subgroups",
    x = "Parental Nativity Status (parborn)",
    y = "Bachelor's Degree or Higher Rate"
  )
```

---

## 6. Life-Course Demographics & Multi-Decade Generational Trends (`age`, `childs`, `year`)

Tracking multi-decade trajectories of degree completion across survey years (1972–2024) by age cohorts (`age`) and fertility (`childs`).

```{r graphic-6}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 6: Longitudinal Trends in Bachelor's+ Attainment (1972–2024) Stratified by Age Groups (age) and Number of Children (childs). Source: GSS (1972–2024)."

df_trend <- df |> 
  filter(!is.na(year_num), !is.na(age_num), !is.na(childs_num), !is.na(educ_cat)) |> 
  mutate(
    age_group = case_when(
      age_num < 35 ~ "18–34 Years",
      age_num <= 55 ~ "35–55 Years",
      TRUE ~ "56+ Years"
    ),
    childs_group = if_else(childs_num >= 3, "3+ Children", "<3 Children")
  ) |> 
  group_by(year_num, age_group, childs_group) |> 
  summarize(
    bachelors_pct = mean(educ_cat %in% c("Bachelor Degree", "Graduate Degree")),
    .groups = "drop"
  )

df_trend |> 
  ggplot(aes(x = year_num, y = bachelors_pct, color = age_group, linetype = childs_group)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("18–34 Years" = "#2563eb", "35–55 Years" = "#7c3aed", "56+ Years" = "#ec4899"), name = "Age Group") +
  scale_linetype_manual(values = c("<3 Children" = "solid", "3+ Children" = "dashed"), name = "Fertility") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Multi-Decade Evolution of College Attainment by Age (age) & Children (childs)",
    subtitle = "Key Takeaway: Steady upward secular expansion from 1972 (15%) to 2024 (38%+), with lower fertility correlating with higher degree completion",
    x = "Survey Administration Year (year)",
    y = "Bachelor's Degree or Higher Rate"
  )
```

---

## 7. Data Coverage & Missingness Spotlight (`agekdbrn`)

Visualizing data availability for `agekdbrn` (age when first child was born) compared to other covariates across GSS survey decades.

```{r graphic-7}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 7: Data Non-Response / Missingness Comparison Across Covariates (Highlighting agekdbrn). Source: GSS (1972–2024)."

df_missing <- tibble(
  Variable = c("educ", "sex", "race", "age", "childs", "sibs", "family16", "res16", "maeduc", "born", "parborn", "incom16", "paeduc", "agekdbrn"),
  Missing_Pct = c(
    mean(is.na(df$educ_cat)),
    mean(is.na(df$sex_clean)),
    mean(is.na(df$race_clean)),
    mean(is.na(df$age_num)),
    mean(is.na(df$childs_num)),
    mean(is.na(df$sibs_num)),
    mean(is.na(df$family16_clean)),
    mean(is.na(df$res16_clean)),
    mean(is.na(df$maeduc_years)),
    mean(is.na(df$born_clean)),
    mean(is.na(df$parborn_clean)),
    mean(is.na(df$incom16_num)),
    mean(is.na(df$paeduc_years)),
    mean(is.na(df$agekdbrn_num))
  )
)

df_missing |> 
  ggplot(aes(x = Missing_Pct, y = reorder(Variable, Missing_Pct), fill = (Variable == "agekdbrn"))) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("FALSE" = "#2563eb", "TRUE" = "#ec4899")) +
  scale_x_continuous(labels = percent_format(), limits = c(0, 0.7)) +
  labs(
    title = "Covariate Non-Response & Missingness Diagnostic (Highlighting agekdbrn)",
    subtitle = "Key Takeaway: agekdbrn exhibits 62.1% missingness/inapplicability, justifying its exclusion from the primary predictive model",
    x = "Percentage of Missing / Inapplicable Observations",
    y = "GSS Covariate Name"
  ) +
  theme(legend.position = "none")
```
