---
title: "Exploratory Data Analysis"
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 9
    fig-height: 5.5
---

::: {.hero-banner}
# Exploratory Data Analysis (EDA) {.unlisted}
<p class="hero-subtitle">Visualizing Interdependencies Across Macroeconomic Growth Drivers & Higher Education Outcomes</p>
:::

In this section, we examine empirical relationships in our macroeconomic panel dataset through 6 distinct visualizations. All data operations strictly follow modern tidyverse R standards using the native R pipe (`|>`).

```{r setup}
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

# Load macro panel dataset
df <- readRDS("data/macro_education_panel.rds")

# Cohesive modern blue-purple-pink ggplot theme
theme_custom <- function() {
  theme_minimal(base_family = "sans") +
    theme(
      plot.title = element_text(face = "bold", size = 13, color = "#0f172a"),
      plot.subtitle = element_text(size = 10, color = "#475569", margin = margin(b = 10)),
      axis.title = element_text(face = "bold", size = 10, color = "#334155"),
      axis.text = element_text(size = 9, color = "#475569"),
      legend.title = element_text(face = "bold", size = 9),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#f1f5f9"),
      strip.background = element_rect(fill = "#1e1b4b", color = NA),
      strip.text = element_text(face = "bold", color = "#ffffff", size = 10)
    )
}
theme_set(theme_custom())

# Brand Color Palette
palette_main <- c("Northeast" = "#2563eb", "Midwest" = "#7c3aed", "South" = "#ec4899", "West" = "#06b6d4")
palette_tiers <- c("Low (<1.0%)" = "#2563eb", "Moderate (1.0-3.0%)" = "#7c3aed", "High (3.0%+)" = "#ec4899")
palette_edu <- c("High School or Less" = "#3b82f6", "Some College" = "#8b5cf6", "Bachelor's+" = "#ec4899")
```

---

## 1. Median Household Income Growth vs. Bachelor's Degree Attainment

Evaluating how real median household income expansion tracks with 4-year college degree completion rates across US state economies.

```{r graphic-1}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 1: Co-movement between annual state median income growth and college completion rates (2004–2024). Source: U.S. Census Bureau (Tables A-1, A-2)."

df |> 
  ggplot(aes(x = income_growth_pct, y = bachelors_plus_pct, color = region)) +
  geom_point(alpha = 0.55, size = 2.2) +
  geom_smooth(method = "loess", color = "#0f172a", fill = "#c084fc", linewidth = 1.3, se = TRUE) +
  scale_color_manual(values = palette_main, name = "Geographic Region") +
  labs(
    title = "Median Household Income Growth vs. Bachelor's Degree Attainment",
    subtitle = "Key Takeaway: State median income growth exhibits strong positive co-movement with bachelor's degree completion rates",
    x = "Annual Real Median Household Income Growth (%)",
    y = "Bachelor's Degree or Higher Attainment (%)"
  )
```

---

## 2. Real Income Growth Distributions Across Educational Attainment Tiers

Examining how annual state median income growth varies across defined educational attainment brackets (*High School or Less*, *Some College*, *Bachelor's+*).

```{r graphic-2}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2: Distribution of real median income growth across educational attainment tiers. Source: U.S. Census Bureau (Table A-2)."

df |> 
  ggplot(aes(x = education_tier, y = income_growth_pct, fill = education_tier)) +
  geom_boxplot(alpha = 0.85, outlier.color = "#ec4899", outlier.alpha = 0.6) +
  scale_fill_manual(values = palette_edu, name = "Education Tier") +
  labs(
    title = "Real Income Growth Stratified by Educational Attainment Tiers",
    subtitle = "Key Takeaway: States in the Bachelor's+ tier demonstrate higher median income growth and lower downside economic dispersion",
    x = "State Educational Attainment Tier",
    y = "Annual Real Median Household Income Growth (%)"
  )
```

---

## 3. Regional GDP Growth Dynamics & High School Completion Over Time

Tracking annual economic expansion ($\Delta\text{GDP}$) overlaid against average high school completion rates across 4 US Census regions over two decades.

```{r graphic-3}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3: Heatmap of annual GDP growth overlaid with mean regional high school completion rates. Source: Macrotrends & U.S. Census Bureau (Table A-1)."

df_heat <- df |> 
  group_by(region, year) |> 
  summarize(
    mean_gdp_growth = mean(gdp_growth_pct),
    mean_hs_pct = mean(hs_completion_pct),
    .groups = "drop"
  )

df_heat |> 
  ggplot(aes(x = factor(year), y = region, fill = mean_gdp_growth)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", mean_hs_pct)), color = "white", size = 2.8, fontface = "bold") +
  scale_fill_gradientn(
    colors = c("#0f172a", "#1e1b4b", "#7c3aed", "#ec4899", "#f472b6"),
    name = "Mean GDP Growth (%)"
  ) +
  labs(
    title = "Temporal Heatmap of Regional GDP Growth & High School Completion",
    subtitle = "Key Takeaway: Macro contractions affect GDP growth uniformly, while high school completion rates maintain steady structural growth",
    x = "Calendar Year",
    y = "Census Region"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 4. Multi-Decade Macroeconomic Predictors vs. Educational Attainment

Tracking multi-decade trajectories of macroeconomic growth drivers ($\Delta\text{Income}$, $\Delta\text{GDP}$, $\Delta\text{Edu Investment}$) alongside college degree completion trends.

```{r graphic-4}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 4: Faceted multi-decade tracking of annual national state means (2004–2024). Source: U.S. Census Bureau, Macrotrends & Simulated Metrics."

df_ts <- df |> 
  group_by(year) |> 
  summarize(
    `Income Growth (%)` = mean(income_growth_pct),
    `GDP Growth (%)` = mean(gdp_growth_pct),
    `Education Investment Growth (%)` = mean(edu_investment_pct),
    `Bachelor's+ Rate (%)` = mean(bachelors_plus_pct),
    .groups = "drop"
  ) |> 
  pivot_longer(cols = -year, names_to = "Metric", values_to = "Value")

df_ts |> 
  ggplot(aes(x = year, y = Value, color = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.0) +
  facet_wrap(~ Metric, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Income Growth (%)" = "#2563eb",
    "GDP Growth (%)" = "#7c3aed",
    "Education Investment Growth (%)" = "#ec4899",
    "Bachelor's+ Rate (%)" = "#06b6d4"
  )) +
  labs(
    title = "Multi-Decade Trajectories of Macroeconomic Predictors & Attainment",
    subtitle = "Key Takeaway: Public education investment and household income growth act as leading indicators for degree attainment",
    x = "Year",
    y = "Indicator Value"
  ) +
  theme(legend.position = "none")
```

---

## 5. Real Income Growth Spread Across Degree Attainment Regimes

Displaying overlapping empirical probability distributions of annual state median income growth, stratified by whether degree attainment is above or below the national median.

```{r graphic-5}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 5: Overlapping probability density of real income growth by attainment regime. Source: U.S. Census Bureau (Tables A-1, A-2)."

df_density <- df |> 
  mutate(
    degree_regime = if_else(
      bachelors_plus_pct >= median(bachelors_plus_pct), 
      "High Attainment (Above Median)", 
      "Low Attainment (Below Median)"
    )
  )

df_density |> 
  ggplot(aes(x = income_growth_pct, fill = degree_regime)) +
  geom_density(alpha = 0.6, color = "white", linewidth = 0.8) +
  scale_fill_manual(values = c(
    "High Attainment (Above Median)" = "#7c3aed", 
    "Low Attainment (Below Median)" = "#ec4899"
  )) +
  labs(
    title = "Density Spread of Real Income Growth Across Attainment Regimes",
    subtitle = "Key Takeaway: States with high college attainment exhibit a rightward shift in income growth and greater economic resilience",
    x = "Annual Real Median Household Income Growth (%)",
    y = "Empirical Probability Density",
    fill = "College Attainment Regime"
  )
```

---

## 6. Higher Education Investment Growth vs. Degree Attainment Across Regions

Evaluating the relationship between state higher education investment growth ($\Delta\text{Edu Investment}$) and college completion rates broken down across US Census Regions.

```{r graphic-6}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 6: Regional faceted linear fits comparing higher education investment and degree completion. Source: U.S. Census Bureau & Simulated Investment Metrics."

df |> 
  ggplot(aes(x = edu_investment_pct, y = bachelors_plus_pct, color = income_tier)) +
  geom_point(alpha = 0.6, size = 2.2) +
  geom_smooth(method = "lm", color = "#0f172a", se = FALSE, linetype = "dashed", linewidth = 1) +
  facet_wrap(~ region, ncol = 2) +
  scale_color_manual(values = palette_tiers, name = "Income Growth Tier") +
  labs(
    title = "Education Investment Growth vs. Degree Completion Across Regions",
    subtitle = "Key Takeaway: Public investment in higher education combined with high income tiers correlates positively with attainment in all regions",
    x = "Annual State Higher Education Investment Growth (%)",
    y = "Bachelor's Degree or Higher Attainment (%)"
  )
```
