---
title: "Exploratory Data Analysis"
subtitle: "Empirical Patterns in Macro-Economic Fluctuations & Educational Outcomes"
format:
  html:
    toc: true
    code-fold: show
    fig-width: 9
    fig-height: 5.5
---

::: {.hero-banner}
# Exploratory Data Analysis (EDA)
### Visualizing Interdependencies Across Economic Indicators & Educational Attainment
:::

In this section, we examine empirical relationships in our macroeconomic panel data using 6 distinct `ggplot2` visualizations. All data operations strictly follow modern tidyverse R standards using the native R pipe (`|>`).

```{r setup}
#| echo: true
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(viridis)
})

# Load macro panel dataset
df <- readRDS("data/macro_education_panel.rds")

# Modern cohesive ggplot theme
theme_custom <- function() {
  theme_minimal(base_family = "sans") +
    theme(
      plot.title = element_text(face = "bold", size = 13, color = "#0f172a"),
      plot.subtitle = element_text(size = 10, color = "#475569", margin = margin(b = 10)),
      axis.title = element_text(face = "bold", size = 10, color = "#334155"),
      axis.text = element_text(size = 9, color = "#475569"),
      legend.title = element_text(face = "bold", size = 9),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "#1e293b", color = NA),
      strip.text = element_text(face = "bold", color = "#ffffff", size = 10)
    )
}
theme_set(theme_custom())
```

---

## Graphic 1: Scatter Plot with Smooth Trendline
### $\Delta\text{Median Income}$ vs. $\% \text{Bachelor's Degree}+$

This scatter plot evaluates whether rising local household wealth ($\Delta\text{Median Income}$) tracks with higher 4-year degree completion percentages across states.

```{r graphic-1}
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Figure 2.1: Median Income Growth vs. Bachelor's Degree Attainment Rate"

df |> 
  ggplot(aes(x = income_growth_pct, y = bachelors_plus_pct, color = region)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "loess", color = "#0f172a", fill = "#cbd5e1", linewidth = 1.2, se = TRUE) +
  scale_color_brewer(palette = "Set1", name = "Geographic Region") +
  labs(
    title = "Figure 2.1: Median Income Growth vs. Bachelor's Degree Attainment",
    subtitle = "Assessing linear and non-linear trend alignment between state income growth and higher education",
    x = "Annual Median Household Income Growth (%)",
    y = "Bachelor's Degree or Higher Attainment (%)"
  )
```

::: {.callout-note}
### EDA Insight: Scatter Plot Analysis
*Insert analysis here.* The non-parametric LOESS trendline illustrates the baseline positive co-movement between median household income expansion and college graduation rates across state economies.
:::

---

## Graphic 2: Categorical Box Plot
### Unemployment Rate Distribution Stratified by Educational Attainment Tiers

This box plot examines how state unemployment rates vary across defined educational attainment tiers (*High School or Less*, *Some College*, *Bachelor's+*).

```{r graphic-2}
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Figure 2.2: State Unemployment Distributions Across Educational Attainment Tiers"

df |> 
  ggplot(aes(x = education_tier, y = unemployment_rate, fill = education_tier)) +
  geom_boxplot(alpha = 0.85, outlier.color = "#e11d48", outlier.alpha = 0.6) +
  scale_fill_viridis_d(option = "mako", name = "Education Tier") +
  labs(
    title = "Figure 2.2: Unemployment Volatility Stratified by Education Tiers",
    subtitle = "Comparing dispersion and median unemployment across educational attainment brackets",
    x = "State Educational Attainment Tier",
    y = "Annual Civilian Unemployment Rate (%)"
  )
```

::: {.callout-tip}
### EDA Insight: Box Plot Analysis
*Insert analysis here.* States situated in the *High School or Less* tier consistently exhibit higher median unemployment rates and wider interquartile ranges during macroeconomic shocks.
:::

---

## Graphic 3: Heatmap / Gradient Map
### Regional GDP Shifts vs. High School Completion Rates Over Time

This spatial-temporal heatmap tracks annual economic growth ($\Delta\text{GDP}$) overlaid against average high school completion rates across 4 US census regions over two decades.

```{r graphic-3}
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Figure 2.3: Heatmap of Regional GDP Growth & High School Completion Over Time"

df_heat <- df |> 
  group_by(region, year) |> 
  summarize(
    mean_gdp_growth = mean(gdp_growth_pct),
    mean_hs_pct = mean(hs_completion_pct),
    .groups = "drop"
  )

df_heat |> 
  ggplot(aes(x = factor(year), y = region, fill = mean_gdp_growth)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.1f%%", mean_hs_pct)), color = "white", size = 2.8, fontface = "bold") +
  scale_fill_viridis_c(option = "magma", name = "Mean GDP Growth (%)") +
  labs(
    title = "Figure 2.3: Temporal Heatmap of Regional GDP Growth",
    subtitle = "Overlaying GDP growth gradients with average High School Completion Rate text labels",
    x = "Calendar Year",
    y = "Census Region"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note}
### EDA Insight: Heatmap Analysis
*Insert analysis here.* Recessions in 2008–2009 and 2020 are visibly captured by darker tiles, while high school completion rates maintain steady upward trends.
:::

---

## Graphic 4: Faceted Time-Series Line Chart
### Multi-Decade Macroeconomic Indicators vs. Educational Attainment

This faceted time-series line chart tracks the multi-decade trajectory of macro indicators ($\Delta\text{GDP}$, $\Delta\text{Income}$, Unemployment) alongside educational completion trends.

```{r graphic-4}
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Figure 2.4: Faceted Multi-Decade Macroeconomic & Educational Time-Series Trends"

df_ts <- df |> 
  group_by(year) |> 
  summarize(
    `GDP Growth (%)` = mean(gdp_growth_pct),
    `Income Growth (%)` = mean(income_growth_pct),
    `Unemployment Rate (%)` = mean(unemployment_rate),
    `Bachelor's+ Rate (%)` = mean(bachelors_plus_pct),
    .groups = "drop"
  ) |> 
  pivot_longer(cols = -year, names_to = "Metric", values_to = "Value")

df_ts |> 
  ggplot(aes(x = year, y = Value, color = Metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ Metric, scales = "free_y", ncol = 2) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Figure 2.4: National Macroeconomic Indicators vs. Higher Education Trends",
    subtitle = "Faceted multi-decade tracking of annual means (2004–2024)",
    x = "Year",
    y = "Indicator Value",
    color = "Macro Indicator"
  ) +
  theme(legend.position = "none")
```

::: {.callout-tip}
### EDA Insight: Time-Series Analysis
*Insert analysis here.* Comparative time-series dynamics show counter-cyclical surges in college completion following macro downturns.
:::

---

## Graphic 5: Stratified Density Plot
### Spread of Annual GDP Growth Stratified by High vs. Low College Enrollment Regimes

This density plot displays overlapping distributions of annual state GDP growth, stratified by whether a state's college degree attainment rate is above or below the national median.

```{r graphic-5}
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Figure 2.5: Density Distribution of GDP Growth by College Attainment Regimes"

df_density <- df |> 
  mutate(
    degree_regime = if_else(
      bachelors_plus_pct >= median(bachelors_plus_pct), 
      "High Attainment (Above Median)", 
      "Low Attainment (Below Median)"
    )
  )

df_density |> 
  ggplot(aes(x = gdp_growth_pct, fill = degree_regime)) +
  geom_density(alpha = 0.55, color = "white", linewidth = 0.8) +
  scale_fill_manual(values = c("High Attainment (Above Median)" = "#0284c7", "Low Attainment (Below Median)" = "#f59e0b")) +
  labs(
    title = "Figure 2.5: Density Spread of GDP Growth Across Attainment Regimes",
    subtitle = "Comparing economic growth volatility between high and low degree attainment states",
    x = "Annual GDP Growth (%)",
    y = "Empirical Probability Density",
    fill = "College Attainment Regime"
  )
```

::: {.callout-note}
### EDA Insight: Density Plot Analysis
*Insert analysis here.* High attainment states exhibit a tighter GDP growth distribution, suggesting greater macroeconomic stability.
:::

---

## Graphic 6: Facet Grid Scatter Plot
### Unemployment vs. Degree Completion Across Geographic Regions

This 4-way faceted scatter grid isolates the correlation between state unemployment rates and college degree attainment broken down across US Geographic Regions.

```{r graphic-6}
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Figure 2.6: Unemployment vs. Degree Attainment Faceted by US Geographic Region"

df |> 
  ggplot(aes(x = unemployment_rate, y = bachelors_plus_pct, color = unemployment_tier)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", color = "#0f172a", se = FALSE, linetype = "dashed") +
  facet_wrap(~ region, ncol = 2) +
  scale_color_brewer(palette = "Dark2", name = "Unemployment Regime") +
  labs(
    title = "Figure 2.6: Regional Heterogeneity in Unemployment vs. Degree Attainment",
    subtitle = "Faceted linear fits capturing structural differences across census regions",
    x = "Annual Civilian Unemployment Rate (%)",
    y = "Bachelor's Degree or Higher Attainment (%)"
  )
```

::: {.callout-tip}
### EDA Insight: Regional Grid Analysis
*Insert analysis here.* Regional variations reveal distinct baseline levels of higher education attainment in the Northeast and West relative to the Midwest and South.
:::
