---
title: "Exploratory Data Analysis"
subtitle: "Empirical Patterns in Macro-Economic Dynamics & Educational Attainment"
format:
  html:
    toc: true
    fig-width: 9
    fig-height: 5.5
---

::: {.hero-banner}
# Exploratory Data Analysis (EDA)
### Visualizing Interdependencies Across Macroeconomic Growth Drivers & Higher Education Outcomes
:::

In this section, we examine empirical relationships in our macroeconomic panel dataset through 6 distinct visualizations. All data operations strictly follow modern tidyverse R standards using the native R pipe (`|>`).

```{r setup}
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

# Load macro panel dataset
df <- readRDS("data/macro_education_panel.rds")

# Cohesive modern blue-purple-pink ggplot theme
theme_custom <- function() {
  theme_minimal(base_family = "sans") +
    theme(
      plot.title = element_text(face = "bold", size = 13, color = "#0f172a"),
      plot.subtitle = element_text(size = 10, color = "#475569", margin = margin(b = 10)),
      axis.title = element_text(face = "bold", size = 10, color = "#334155"),
      axis.text = element_text(size = 9, color = "#475569"),
      legend.title = element_text(face = "bold", size = 9),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#f1f5f9"),
      strip.background = element_rect(fill = "#1e1b4b", color = NA),
      strip.text = element_text(face = "bold", color = "#ffffff", size = 10)
    )
}
theme_set(theme_custom())

# Brand Color Palette
palette_main <- c("Northeast" = "#2563eb", "Midwest" = "#7c3aed", "South" = "#ec4899", "West" = "#06b6d4")
palette_tiers <- c("Low (<1.0%)" = "#2563eb", "Moderate (1.0-3.0%)" = "#7c3aed", "High (3.0%+)" = "#ec4899")
palette_edu <- c("High School or Less" = "#3b82f6", "Some College" = "#8b5cf6", "Bachelor's+" = "#ec4899")
```

---

## Graphic 1: Scatter Plot with Smooth Trendline
### $\Delta\text{Median Income}$ vs. $\% \text{Bachelor's Degree}+$

This scatter plot evaluates how local household income growth ($\Delta\text{Median Income}$) tracks with 4-year college degree completion rates across US state economies.

```{r graphic-1}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2.1: Median Income Growth vs. Bachelor's Degree Attainment Rate"

df |> 
  ggplot(aes(x = income_growth_pct, y = bachelors_plus_pct, color = region)) +
  geom_point(alpha = 0.55, size = 2.2) +
  geom_smooth(method = "loess", color = "#0f172a", fill = "#c084fc", linewidth = 1.3, se = TRUE) +
  scale_color_manual(values = palette_main, name = "Geographic Region") +
  labs(
    title = "Figure 2.1: Median Household Income Growth vs. Bachelor's Degree Attainment",
    subtitle = "Evaluating linear and non-linear trend alignment between state income growth and higher education rates",
    x = "Annual Real Median Household Income Growth (%)",
    y = "Bachelor's Degree or Higher Attainment (%)"
  )
```

::: {.callout-note}
### EDA Insight: Scatter Plot Analysis
The non-parametric LOESS trendline illustrates a strong positive co-movement between median household income expansion and college completion rates across state economies. States experiencing sustained real median income growth consistently achieve higher levels of higher educational attainment.
:::

---

## Graphic 2: Categorical Box Plot
### Real Income Growth Stratified by Educational Attainment Tiers

This box plot examines how annual median income growth varies across defined educational attainment tiers (*High School or Less*, *Some College*, *Bachelor's+*).

```{r graphic-2}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2.2: State Income Growth Distributions Across Educational Attainment Tiers"

df |> 
  ggplot(aes(x = education_tier, y = income_growth_pct, fill = education_tier)) +
  geom_boxplot(alpha = 0.85, outlier.color = "#ec4899", outlier.alpha = 0.6) +
  scale_fill_manual(values = palette_edu, name = "Education Tier") +
  labs(
    title = "Figure 2.2: Real Income Growth Distribution Stratified by Education Tiers",
    subtitle = "Comparing median growth and interquartile dispersion across state educational attainment brackets",
    x = "State Educational Attainment Tier",
    y = "Annual Real Median Household Income Growth (%)"
  )
```

::: {.callout-tip}
### EDA Insight: Box Plot Analysis
States situated in the *Bachelor's+* attainment tier demonstrate significantly higher median real income growth and lower downside dispersion, highlighting the protective buffer provided by a highly educated workforce.
:::

---

## Graphic 3: Heatmap / Gradient Map
### Regional GDP Growth Shifts & High School Completion Rates Over Time

This spatial-temporal heatmap tracks annual economic growth ($\Delta\text{GDP}$) overlaid against average high school completion rates across 4 US Census regions over two decades.

```{r graphic-3}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2.3: Heatmap of Regional GDP Growth & High School Completion Over Time"

df_heat <- df |> 
  group_by(region, year) |> 
  summarize(
    mean_gdp_growth = mean(gdp_growth_pct),
    mean_hs_pct = mean(hs_completion_pct),
    .groups = "drop"
  )

df_heat |> 
  ggplot(aes(x = factor(year), y = region, fill = mean_gdp_growth)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", mean_hs_pct)), color = "white", size = 2.8, fontface = "bold") +
  scale_fill_gradientn(
    colors = c("#0f172a", "#1e1b4b", "#7c3aed", "#ec4899", "#f472b6"),
    name = "Mean GDP Growth (%)"
  ) +
  labs(
    title = "Figure 2.3: Temporal Heatmap of Regional GDP Growth",
    subtitle = "Overlaying regional GDP growth gradients with average High School Completion Rate text labels",
    x = "Calendar Year",
    y = "Census Region"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note}
### EDA Insight: Heatmap Analysis
Macroeconomic contraction periods (e.g., 2008–2009 and 2020) appear as deep violet/dark tiles across all regions, while underlying high school completion rates maintain steady structural growth over time.
:::

---

## Graphic 4: Faceted Time-Series Line Chart
### Multi-Decade Macroeconomic Growth Drivers vs. Educational Attainment

This faceted time-series line chart tracks the multi-decade trajectory of macro growth drivers ($\Delta\text{Income}$, $\Delta\text{GDP}$, $\Delta\text{Edu Investment}$) alongside college degree completion trends.

```{r graphic-4}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2.4: Faceted Multi-Decade Macroeconomic & Educational Time-Series Trends"

df_ts <- df |> 
  group_by(year) |> 
  summarize(
    `Income Growth (%)` = mean(income_growth_pct),
    `GDP Growth (%)` = mean(gdp_growth_pct),
    `Education Investment Growth (%)` = mean(edu_investment_pct),
    `Bachelor's+ Rate (%)` = mean(bachelors_plus_pct),
    .groups = "drop"
  ) |> 
  pivot_longer(cols = -year, names_to = "Metric", values_to = "Value")

df_ts |> 
  ggplot(aes(x = year, y = Value, color = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.0) +
  facet_wrap(~ Metric, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Income Growth (%)" = "#2563eb",
    "GDP Growth (%)" = "#7c3aed",
    "Education Investment Growth (%)" = "#ec4899",
    "Bachelor's+ Rate (%)" = "#06b6d4"
  )) +
  labs(
    title = "Figure 2.4: Macroeconomic Predictors vs. Higher Education Trends",
    subtitle = "Faceted multi-decade tracking of national annual state means (2004–2024)",
    x = "Year",
    y = "Indicator Value"
  ) +
  theme(legend.position = "none")
```

::: {.callout-tip}
### EDA Insight: Time-Series Analysis
State higher education investment growth and median household income expansion show strong co-movement, serving as persistent leading indicators for state degree attainment trends.
:::

---

## Graphic 5: Stratified Density Plot
### Real Income Growth Spread Stratified by Degree Attainment Regimes

This density plot displays overlapping empirical probability distributions of annual state median income growth, stratified by whether a state's college degree attainment rate is above or below the national median.

```{r graphic-5}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2.5: Density Spread of Real Income Growth Across Attainment Regimes"

df_density <- df |> 
  mutate(
    degree_regime = if_else(
      bachelors_plus_pct >= median(bachelors_plus_pct), 
      "High Attainment (Above Median)", 
      "Low Attainment (Below Median)"
    )
  )

df_density |> 
  ggplot(aes(x = income_growth_pct, fill = degree_regime)) +
  geom_density(alpha = 0.6, color = "white", linewidth = 0.8) +
  scale_fill_manual(values = c(
    "High Attainment (Above Median)" = "#7c3aed", 
    "Low Attainment (Below Median)" = "#ec4899"
  )) +
  labs(
    title = "Figure 2.5: Density Spread of Real Income Growth Across Attainment Regimes",
    subtitle = "Comparing economic growth distribution between high and low degree attainment states",
    x = "Annual Real Median Household Income Growth (%)",
    y = "Empirical Probability Density",
    fill = "College Attainment Regime"
  )
```

::: {.callout-note}
### EDA Insight: Density Plot Analysis
States in the high attainment regime demonstrate a rightward shift in their income growth distribution, confirming higher mean growth and greater economic resilience compared to low attainment states.
:::

---

## Graphic 6: Facet Grid Scatter Plot
### Education Investment Growth vs. Degree Attainment Across Regions

This 4-way faceted scatter grid evaluates the relationship between state higher education investment growth ($\Delta\text{Edu Investment}$) and college completion rates broken down across US Census Regions.

```{r graphic-6}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2.6: Higher Education Investment vs. Degree Attainment Faceted by Region"

df |> 
  ggplot(aes(x = edu_investment_pct, y = bachelors_plus_pct, color = income_tier)) +
  geom_point(alpha = 0.6, size = 2.2) +
  geom_smooth(method = "lm", color = "#0f172a", se = FALSE, linetype = "dashed", linewidth = 1) +
  facet_wrap(~ region, ncol = 2) +
  scale_color_manual(values = palette_tiers, name = "Income Growth Tier") +
  labs(
    title = "Figure 2.6: Regional Heterogeneity in Education Investment vs. Degree Completion",
    subtitle = "Faceted linear fits capturing structural differences across US Census Regions",
    x = "Annual State Higher Education Investment Growth (%)",
    y = "Bachelor's Degree or Higher Attainment (%)"
  )
```

::: {.callout-tip}
### EDA Insight: Regional Grid Analysis
Across all four regions, increased public investment in higher education combined with high income growth tiers correlates positively with elevated bachelor's degree attainment rates.
:::
