::: {.hero-banner}
# Predictive Drivers of Educational Attainment {.unlisted}
<p class="hero-subtitle">How parental background, childhood economic status, family structure, geography, and birth cohorts shape educational outcomes in the United States.</p>
:::

::: {.overview-graphic-card}
<div class="overview-graphic-header">
  <span class="overview-graphic-badge">Interactive Overview</span>
  <h3 style="margin: 0; font-size: 1.2rem; font-weight: 700; color: #0f172a;">50-Year Trajectories of Educational Mobility (1972–2024)</h3>
</div>

<p style="font-size: 0.95rem; color: #475569; margin-bottom: 1.25rem;">
  Explore how college graduation rates evolved over 50 years across childhood income levels. Even on a still frame, each line traces the rising educational trajectory of an income group from the 1970s to the 2020s. Hover over any point to inspect detailed metrics, or click legend entries to isolate specific childhood income tiers.
</p>

```{r hero-animated-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Interactive Figure: 50-Year Trajectories of Bachelor's degree or higher completion rates (%) across survey decades (1970s–2020s) stratified by childhood income tier. Data Source: General Social Survey."

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(plotly)
})

df <- readRDS("data/gss_cleaned.rds")

df_traj <- df |> 
  filter(!is.na(year_num), !is.na(paeduc_years), !is.na(educ_cat), !is.na(incom16_cat)) |> 
  mutate(
    decade = paste0(floor(year_num / 10) * 10, "s"),
    incom16_cat = factor(
      incom16_cat,
      levels = c("FAR BELOW AVERAGE", "BELOW AVERAGE", "AVERAGE", "ABOVE AVERAGE", "FAR ABOVE AVERAGE"),
      labels = c("Far Below Average", "Below Average", "Average", "Above Average", "Far Above Average")
    )
  ) |> 
  group_by(decade, incom16_cat) |> 
  summarize(
    bachelors_pct = round(mean(educ_cat %in% c("Bachelor Degree", "Graduate Degree")) * 100, 1),
    avg_parent_educ = round(mean(paeduc_years, na.rm = TRUE), 1),
    avg_resp_years = round(mean(educ_years, na.rm = TRUE), 1),
    sample_size = n(),
    .groups = "drop"
  )

income_colors <- c(
  "Far Above Average" = "#ef4466",
  "Above Average"     = "#ec4899",
  "Average"           = "#7c3aed",
  "Below Average"     = "#2563eb",
  "Far Below Average" = "#06b6d4"
)

fig <- plot_ly(
  data = df_traj,
  x = ~decade,
  y = ~bachelors_pct,
  color = ~incom16_cat,
  colors = income_colors,
  type = "scatter",
  mode = "lines+markers",
  line = list(width = 3.5),
  marker = list(size = 9, line = list(width = 2, color = "#ffffff")),
  text = ~paste0(
    "<b>Decade: ", decade, "</b><br>",
    "<b>Childhood Income:</b> ", incom16_cat, "<br>",
    "<b>Bachelor's Degree+ Rate:</b> ", bachelors_pct, "%<br>",
    "<b>Avg Parental Education:</b> ", avg_parent_educ, " Years<br>",
    "<b>Avg Respondent Schooling:</b> ", avg_resp_years, " Years<br>",
    "<b>Sample Size:</b> ", format(sample_size, big.mark = ",")
  ),
  hoverinfo = "text"
) |> 
layout(
  xaxis = list(
    title = list(text = "<b>Survey Decade</b>", font = list(size = 12, color = "#0f172a")),
    gridcolor = "#f1f5f9"
  ),
  yaxis = list(
    title = list(text = "<b>Bachelor's Degree or Higher Rate (%)</b>", font = list(size = 12, color = "#0f172a")),
    gridcolor = "#f1f5f9",
    ticksuffix = "%",
    range = c(0, 75)
  ),
  legend = list(
    title = list(text = "<b>Childhood Income Tier</b>"),
    orientation = "h",
    x = 0.02,
    y = -0.22
  ),
  paper_bgcolor = "transparent",
  plot_bgcolor = "#ffffff",
  margin = list(t = 20, b = 70, l = 60, r = 30)
)

fig
```
:::

## 1. The Context

Educational attainment drives socio-economic mobility, career earnings, and health trajectories. While higher education access has expanded significantly over the past five decades, individual educational paths remain strongly tied to childhood family backgrounds and geographic origins.

Quantifying how these early life factors jointly predict educational outcomes helps sociologists and policy planners identify structural barriers and evaluate intergenerational mobility in the United States.

---

## 2. Central Predictive Research Question

To what extent do social, economic, and demographic background factors jointly predict whether an individual achieves a post-secondary degree, such as a Bachelor's degree, in the United States?

---

## 3. Dataset Architecture

This study uses the national **General Social Survey (GSS)** panel spanning **52 years (1972–2024)** with **$N = 75,699$ observations**, tracking individual educational attainment alongside key socio-demographic background variables.

```{r}
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(knitr)
})

# Load clean GSS dataset
df <- readRDS("data/gss_cleaned.rds")

# High-level dataset summary table covering ALL variables
summary_table <- tibble(
  `Variable Name` = c(
    "educ (Outcome)", "paeduc", "maeduc", "incom16", "family16",
    "res16", "sex", "race", "born", "parborn",
    "sibs", "childs", "age", "year", "agekdbrn"
  ),
  `Variable Role & Type` = c(
    "Primary Target (Categorical)", "Covariate (Years)", "Covariate (Years)", 
    "Covariate (Categorical)", "Covariate (Categorical)", "Covariate (Categorical)", 
    "Covariate (Binary)", "Covariate (Categorical)", "Covariate (Binary)", 
    "Covariate (Categorical)", "Covariate (Discrete Count)", "Covariate (Discrete Count)", 
    "Covariate (Continuous Years)", "Covariate (Continuous Year)", "Covariate (Continuous Years)"
  ),
  Description = c(
    "Respondent's Highest Educational Attainment Tier (High School or Less, Some College, Bachelor's, Graduate)",
    "Father's Educational Attainment (Years 0–20)",
    "Mother's Educational Attainment (Years 0–20)",
    "Family Relative Income at Age 16 (Far Below Average to Far Above Average)",
    "Living Arrangement / Family Structure at Age 16 (Both Parents, Single Parent, Stepparent, Other)",
    "Type of Place Lived in at Age 16 (Farm, Town <50k, Suburb, City >250k)",
    "Respondent's Sex (Female, Male)",
    "Respondent's Race (White, Black, Other)",
    "Was Respondent Born in the United States? (Yes, No)",
    "Were Respondent's Parents Born in the United States? (Both U.S., One U.S., Neither U.S.)",
    "Number of Brothers and Sisters / Siblings (0 to 6+)",
    "Number of Children (0 to 8+)",
    "Respondent's Age at Time of Survey (18 to 89)",
    "Survey Administration Year (1972 to 2024)",
    "Age When First Child Was Born"
  ))

summary_table |> 
  kable(caption = "Table 1.1: General Social Survey (GSS) Variable Specification & Data Coverage (1972–2024)")
```

---

## 4. Assumptions & Counters

- **Validity:** The consistency between measured columns and the ideal population table.
  - *Counter:* The GSS variable `educ` measures total schooling years rather than degree credentials. Standard threshold conversions (e.g., 4 years = Bachelor's degree) can misclassify individuals who take extra years to graduate or finish early.

- **Stability:** The assumption that covariate relationships remain constant over time.
  - *Counter:* In data spanning 1972 to 2024, societal changes—such as rising female labor force participation and college enrollment—mean older survey cohorts may exhibit different mobility dynamics than recent cohorts.

- **Representativeness:** The assumption that sample rows represent the target population.
  - *Counter:* While the GSS uses a national probability sample, older cohorts may reflect historical educational structures that differ from current trends.

- **Unconfoundedness:** The assumption of independent treatment assignment.
  - *Counter:* Not applicable, as this study is strictly predictive rather than causal.

---

## 5. Key Predictive Findings

Our predictive models reveal that childhood background factors strongly shape whether an individual earns a college degree. Ranked from **most to least influential**:

1. **Parental Education (Most Influential Factor)**: A parent's own education is the single strongest predictor of a child's degree attainment. Each additional year of schooling completed by parents is associated with a **31.8% to 34.7% increase in the odds** of achieving a Bachelor's degree (adding **+0.33 years** of schooling).
2. **Family Structure**: Growing up with a single parent is associated with a **40.4% to 47.0% decrease in the odds** of earning a Bachelor's degree compared to growing up with both parents (reducing expected schooling by **0.61 years**).
3. **Number of Siblings**: Coming from a larger family is linked to lower college completion rates. Each additional sibling in the household is associated with a **13.2% to 16.0% decrease in the odds** of graduating with a Bachelor's degree (**-0.18 years** per sibling).
4. **Immigrant Parent Background**: Children of foreign-born parents demonstrate high educational mobility, with non-U.S.-born parental nativity associated with a **71.3% to 82.8% increase in the odds** of attaining a Bachelor's degree (**+0.51 years** of completed schooling).
5. **Childhood Household Income**: Higher relative family income during childhood steadily opens doors to higher education. Each step up in childhood income tier is associated with a **13.2% to 21.4% increase in the odds** of completing a Bachelor's degree (**+0.13 years** per income level).
6. **Suburban Growing Environment**: Growing up in a suburban community offers higher college graduation odds compared to a rural or farm environment, associated with a **35.8% to 47.9% increase in the odds** of earning a Bachelor's degree (**+0.25 years** of schooling).

> *For technical statistical parameters, test statistics ($t$-stats and $z$-scores), and formal confidence interval specifications, see the **[DGM & Predictive Model](model.qmd)** tab.*

Overall, our predictive model accurately determines whether an individual completes a college degree for **72.3% of individuals** based solely on these childhood background traits.
