::: {.hero-banner}
# Predictive Model & Verdict {.unlisted}
<p class="hero-subtitle">Statistical Formulations, Parameter Estimation, 95% Confidence Intervals, and Empirical Performance</p>
:::

```{r setup}
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(broom)
  library(nnet)
  library(knitr)
  library(scales)
})

# Load clean modeling dataset and fitted models
m_data <- readRDS("data/gss_model_data.rds")
multi_fit <- readRDS("data/gss_multinomial_model.rds")
ols_fit <- readRDS("data/gss_ols_model.rds")

# Custom ggplot theme with left-aligned titles to prevent text clipping
theme_set(
  theme_minimal(base_family = "sans") +
    theme(
      plot.title.position = "plot",
      plot.title = element_text(face = "bold", size = 12, color = "#0f172a"),
      plot.subtitle = element_text(size = 9, color = "#475569", margin = margin(b = 8)),
      axis.title = element_text(face = "bold", size = 9.5, color = "#334155"),
      axis.text = element_text(size = 8.5, color = "#475569"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#f1f5f9")
    )
)
```

---

## 1. Data Generating Mechanism (DGM) Formulations

To predict individual educational attainment ($Y_{\text{educ}}$) conditioned on childhood background traits, we evaluate two statistical models: a primary **Multinomial Logistic Model** for milestone degree tiers and a **Linear Regression (OLS) Model** for continuous schooling years.

### 1.1 Multinomial Logistic DGM

We model the log-odds of achieving educational tier $k \in \{\text{Some College}, \text{Bachelor Degree}, \text{Graduate Degree}\}$ relative to baseline $K_0 = \text{High School or Less}$:

$$\ln\left(\frac{P(Y_{\text{educ}} = k \mid \mathbf{X})}{P(Y_{\text{educ}} = K_0 \mid \mathbf{X})}\right) = \beta_{k0} + \boldsymbol{\beta}_k^T \mathbf{X}$$

where $\mathbf{X}$ is the vector of childhood socio-demographic covariates:

$$\mathbf{X} = \Big[ \text{parent\_educ}, \text{incom16}, \text{sibs}, \text{childs}, \text{age}, \text{year}, \text{sex}, \text{race}, \text{res16}, \text{family16}, \text{born}, \text{parborn} \Big]^T$$

The predicted probability $P(Y_{\text{educ}} = k \mid \mathbf{X})$ across all 4 tiers is obtained via the soft-max transformation:

$$P(Y_{\text{educ}} = k \mid \mathbf{X}) = \frac{\exp\left(\beta_{k0} + \boldsymbol{\beta}_k^T \mathbf{X}\right)}{1 + \sum_{j=1}^{3} \exp\left(\beta_{j0} + \boldsymbol{\beta}_j^T \mathbf{X}\right)}$$

---

### 1.2 Linear Regression DGM

As a baseline comparison, completed schooling years ($Y_{\text{years}} \in [0, 20]$) are modeled as a linear conditional expectation:

$$Y_{\text{years}} = \beta_0 + \boldsymbol{\beta}^T \mathbf{X} + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2)$$

---

## 2. Model Parameters & 95% Confidence Intervals

> **Note**: Tables 2.1 & 2.2 highlight key predictors. Complete parameter tables across all variables are available in the [Sources](sources.qmd) tab.

### 2.1 Multinomial Logistic Model Parameters

Parameters are exponentiated into Odds Ratios ($\text{OR} = \exp(\beta)$) with 95% Confidence Intervals.

```{r fit-multinomial-table}
#| echo: false
#| message: false
#| warning: false

multi_tidy <- broom::tidy(multi_fit, conf.int = TRUE, exponentiate = TRUE)

verdict_multi_terms <- c(
  "parent_educ_avg",
  "incom16_num",
  "family16_cleanSingle Parent",
  "sibs_num",
  "res16_cleanBIG-CITY SUBURB",
  "parborn_cleanNeither U.S."
)

multi_table <- multi_tidy |> 
  filter(term %in% verdict_multi_terms, y.level == "Bachelor Degree") |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Relative Income (Tier 1-5)",
      term == "sibs_num" ~ "Sibling Count (Per Sibling)",
      term == "family16_cleanSingle Parent" ~ "Single Parent Family (vs Both Parents)",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Residence: Big-City Suburb (vs Farm)",
      term == "parborn_cleanNeither U.S." ~ "Parents Nativity: Neither U.S. Born (vs Both)",
      TRUE ~ term
    ),
    across(c(estimate, conf.low, conf.high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "< 0.001", as.character(round(p.value, 4)))
  ) |> 
  select(
    `Outcome Tier` = y.level,
    `Predictor Variable` = term_clean,
    `Odds Ratio (OR)` = estimate,
    `p-value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  )

multi_table |> 
  kable(caption = "Table 2.1: Multinomial Odds Ratios & 95% Confidence Intervals for Bachelor's Degree Attainment (Ref: High School or Less)")
```

**Interpretation**: An Odds Ratio (OR) > 1 indicates increased likelihood of degree completion, while OR < 1 indicates decreased likelihood. For instance, each additional year of parental education increases Bachelor's degree completion odds by **33.3%** ($\text{OR} = 1.333$).

---

### 2.2 OLS Model Parameters

```{r fit-ols-table}
#| echo: false
#| message: false
#| warning: false

ols_tidy <- broom::tidy(ols_fit, conf.int = TRUE)

verdict_ols_terms <- c(
  "parent_educ_avg",
  "incom16_num",
  "family16_cleanSingle Parent",
  "sibs_num",
  "res16_cleanBIG-CITY SUBURB",
  "parborn_cleanNeither U.S."
)

ols_tidy |> 
  filter(term %in% verdict_ols_terms) |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Relative Income (Tier 1-5)",
      term == "family16_cleanSingle Parent" ~ "Family Structure: Single Parent",
      term == "sibs_num" ~ "Sibling Count (Per Sibling)",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Residence Origin: Big-City Suburb",
      term == "parborn_cleanNeither U.S." ~ "Parents Nativity: Neither U.S. Born",
      TRUE ~ term
    ),
    across(c(estimate, std.error, statistic, conf.low, conf.high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "< 0.001", as.character(round(p.value, 4)))
  ) |> 
  select(
    `Predictor` = term_clean,
    `Estimate (β)` = estimate,
    `p-value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  ) |> 
  kable(caption = "Table 2.2: OLS Linear Regression Parameters & 95% Confidence Intervals for Completed Schooling Years")
```

**Interpretation**: Estimate ($\beta$) measures change in total completed schooling years per 1-unit increase in the predictor. For example, each year of parental education adds **+0.326 years** of completed schooling.

---

## 3. Visualizing Model Uncertainty

### 3.1 Multinomial Odds Ratio Uncertainty

Figure 3.1 displays Odds Ratios ($\text{OR} = \exp(\beta)$) and 95% Confidence Intervals for predicting **Bachelor Degree** and **Graduate Degree** attainment relative to High School baseline.

```{r forest-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3.1: Multinomial Logistic Model Odds Ratios & 95% Confidence Intervals for Bachelor's and Graduate Degrees."

plot_terms <- c("parent_educ_avg", "incom16_num", "sibs_num", "childs_num", "parborn_cleanNeither U.S.", "family16_cleanSingle Parent", "res16_cleanBIG-CITY SUBURB")

multi_tidy |> 
  filter(y.level %in% c("Bachelor Degree", "Graduate Degree"), term %in% plot_terms) |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Income Tier (+1 Level)",
      term == "sibs_num" ~ "Sibling Count (+1 Sibling)",
      term == "childs_num" ~ "Number of Children (+1 Child)",
      term == "parborn_cleanNeither U.S." ~ "Parents Born: Neither U.S. vs Both",
      term == "family16_cleanSingle Parent" ~ "Single Parent vs Both Parents",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Big-City Suburb vs Farm Origin"
    )
  ) |> 
  ggplot(aes(x = estimate, y = reorder(term_clean, estimate), color = y.level, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "#ec4899", linewidth = 0.9) +
  geom_errorbarh(height = 0.3, position = position_dodge(width = 0.5), linewidth = 1.0) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("Bachelor Degree" = "#2563eb", "Graduate Degree" = "#7c3aed"), name = "Outcome Tier") +
  labs(
    title = "Multinomial Model Odds Ratios & 95% Confidence Intervals",
    subtitle = "OR > 1 increases degree odds; OR < 1 decreases odds (Ref: High School or Less)",
    x = "Odds Ratio [OR = exp(β)]",
    y = "Predictor Variable"
  ) +
  theme(
    legend.position = "bottom"
  )
```

---

### 3.2 OLS Regression Uncertainty

Figure 3.2 displays continuous linear parameter estimates ($\beta$) and 95% Confidence Intervals for total completed schooling years ($Y_{\text{years}}$). Dashed line ($\beta = 0$) represents no effect.

```{r ols-forest-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3.2: OLS Parameter Estimates (β) & 95% Confidence Intervals for Completed Schooling Years."

ols_plot_terms <- c(
  "parent_educ_avg",
  "incom16_num",
  "sibs_num",
  "childs_num",
  "parborn_cleanNeither U.S.",
  "family16_cleanSingle Parent",
  "res16_cleanBIG-CITY SUBURB"
)

ols_tidy |> 
  filter(term %in% ols_plot_terms) |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Income Tier (+1 Level)",
      term == "sibs_num" ~ "Sibling Count (+1 Sibling)",
      term == "childs_num" ~ "Number of Children (+1 Child)",
      term == "parborn_cleanNeither U.S." ~ "Parents Born: Neither U.S. vs Both",
      term == "family16_cleanSingle Parent" ~ "Single Parent vs Both Parents",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Big-City Suburb vs Farm Origin"
    ),
    effect_dir = if_else(estimate >= 0, "Positive Effect (+ Years)", "Negative Effect (- Years)"),
    label_text = sprintf("%+.2f yrs", estimate)
  ) |> 
  ggplot(aes(x = estimate, y = reorder(term_clean, estimate), color = effect_dir, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#64748b", linewidth = 0.9) +
  geom_errorbarh(height = 0.25, linewidth = 1.1) +
  geom_point(size = 3.5, shape = 18) +
  geom_text(
    aes(label = label_text),
    vjust = -0.8,
    size = 3.0,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c("Positive Effect (+ Years)" = "#0d9488", "Negative Effect (- Years)" = "#e11d48"),
    name = "Effect Direction"
  ) +
  labs(
    title = "OLS Model Estimates & 95% Confidence Intervals",
    subtitle = "Change in total schooling years per predictor unit (Dashed line β = 0 is null)",
    x = "Effect Size on Schooling Years (β in Years)",
    y = "Predictor Variable"
  ) +
  theme(
    legend.position = "bottom"
  )
```

---

## 4. Practical Profile Predictions

To illustrate model behavior, we evaluate two individual profiles with contrasting childhood backgrounds.

```{r profile-predictions}
#| echo: false
#| message: false
#| warning: false

# Profile 1: Advantaged Profile
profile_a <- tibble(
  parent_educ_avg = 16,
  incom16_num = 5,
  sibs_num = 1,
  childs_num = 0,
  age_num = 30,
  year_num = 2020,
  sex_clean = "FEMALE",
  race_clean = "White",
  res16_clean = "BIG-CITY SUBURB",
  family16_clean = "Both Own Parents",
  born_clean = "YES",
  parborn_clean = "Both U.S."
)

# Profile 2: Disadvantaged Profile
profile_b <- tibble(
  parent_educ_avg = 9,
  incom16_num = 1,
  sibs_num = 4,
  childs_num = 0,
  age_num = 30,
  year_num = 2020,
  sex_clean = "MALE",
  race_clean = "White",
  res16_clean = "FARM",
  family16_clean = "Single Parent",
  born_clean = "YES",
  parborn_clean = "Both U.S."
)

profiles <- bind_rows(profile_a, profile_b)
probs <- predict(multi_fit, newdata = profiles, type = "probs")
ols_years <- predict(ols_fit, newdata = profiles)

profile_summary <- tibble(
  `Model Metric / Outcome` = c(
    "Childhood Background",
    "High School or Less Probability",
    "Some College Probability",
    "Bachelor Degree Probability",
    "Graduate Degree Probability",
    "Expected Total Schooling Years"
  ),
  `Individual A (Advantaged)` = c(
    "College-educated parents, High income, Two parents, Suburban, 1 Sibling",
    paste0(round(probs[1, "High School or Less"] * 100, 1), "%"),
    paste0(round(probs[1, "Some College"] * 100, 1), "%"),
    paste0(round(probs[1, "Bachelor Degree"] * 100, 1), "%"),
    paste0(round(probs[1, "Graduate Degree"] * 100, 1), "%"),
    paste0(round(ols_years[1], 1), " Years")
  ),
  `Individual B (Disadvantaged)` = c(
    "Middle-school parents, Low income, Single parent, Farm origin, 4 Siblings",
    paste0(round(probs[2, "High School or Less"] * 100, 1), "%"),
    paste0(round(probs[2, "Some College"] * 100, 1), "%"),
    paste0(round(probs[2, "Bachelor Degree"] * 100, 1), "%"),
    paste0(round(probs[2, "Graduate Degree"] * 100, 1), "%"),
    paste0(round(ols_years[2], 1), " Years")
  )
)

profile_summary |> 
  kable(caption = "Table 2.3: Model-Predicted Educational Attainment Probabilities & Expected Schooling Years")
```

- **Individual A**: High parental education and economic stability yield a **39.7% chance** of a Bachelor's degree (**67.9% chance** of any college degree) and **16.7 expected schooling years**.
- **Individual B**: Cumulative background challenges reduce Bachelor's degree probability to **5.1%** (**70.4% chance** of stopping at High School) and **11.9 expected schooling years**.

---

## 5. Summary & Empirical Predictor Importance

To rigorously evaluate feature importance across our statistical DGMs, predictors are ranked below by empirical test statistics ($t$-statistics for continuous schooling and $z$-scores for degree tier multinomial logistic regression) alongside 95% confidence intervals and effect sizes.

### 5.1 Predictor Importance Ranking (Most to Least Important)

1. **Parental Educational Background ($t = 84.48, z = 52.54, p < 0.001$)**:
   - **OLS Linear Effect ($\beta$)**: $+0.326$ years of completed schooling per parent year ($95\%\text{ CI: } [0.318, 0.333]$).
   - **Multinomial Odds Ratio**: $\text{OR} = 1.333$ ($95\%\text{ CI: } [1.318, 1.347]$) for a Bachelor's degree; $\text{OR} = 1.405$ ($95\%\text{ CI: } [1.388, 1.423]$) for a Graduate degree.
   - **Statistical Verdict**: The single most powerful background predictor across both linear and multinomial specifications.

2. **Family Structure / Single-Parent Household ($t = -18.18, z = -19.27, p < 0.001$)**:
   - **OLS Linear Effect ($\beta$)**: $-0.607$ schooling years ($95\%\text{ CI: } [-0.672, -0.541]$) compared to two-parent households.
   - **Multinomial Odds Ratio**: $\text{OR} = 0.562$ ($95\%\text{ CI: } [0.530, 0.596]$) for a Bachelor's degree.
   - **Statistical Verdict**: Substantial negative penalty associated with single-parent childhood environments.

3. **Family Sibling Size ($t = -27.77, z = -18.84, p < 0.001$)**:
   - **OLS Linear Effect ($\beta$)**: $-0.182$ schooling years per additional sibling ($95\%\text{ CI: } [-0.195, -0.170]$).
   - **Multinomial Odds Ratio**: $\text{OR} = 0.854$ ($95\%\text{ CI: } [0.840, 0.868]$) per sibling for a Bachelor's degree.
   - **Statistical Verdict**: Significant dilution of per-child educational resources in larger families.

4. **Parental Immigrant Status ($t = 9.30, z = 34.18, p < 0.001$)**:
   - **OLS Linear Effect ($\beta$)**: $+0.508$ schooling years ($95\%\text{ CI: } [0.401, 0.615]$) for non-U.S. born parents vs. U.S. born parents.
   - **Multinomial Odds Ratio**: $\text{OR} = 1.770$ ($95\%\text{ CI: } [1.713, 1.828]$) for a Bachelor's degree.
   - **Statistical Verdict**: Strong generational upward mobility demonstrated by children of foreign-born parents.

5. **Childhood Relative Income ($t = 9.17, z = 9.00, p < 0.001$)**:
   - **OLS Linear Effect ($\beta$)**: $+0.129$ schooling years per income level ($95\%\text{ CI: } [0.101, 0.157]$).
   - **Multinomial Odds Ratio**: $\text{OR} = 1.172$ ($95\%\text{ CI: } [1.132, 1.214]$) per tier for a Bachelor's degree.
   - **Statistical Verdict**: Consistent, positive linear returns to childhood relative economic standing.

6. **Suburban Origin Geography ($t = 5.82, z = 16.31, p < 0.001$)**:
   - **OLS Linear Effect ($\beta$)**: $+0.246$ schooling years ($95\%\text{ CI: } [0.163, 0.329]$) for big-city suburbs vs. farm origin.
   - **Multinomial Odds Ratio**: $\text{OR} = 1.420$ ($95\%\text{ CI: } [1.358, 1.479]$) for a Bachelor's degree.
   - **Statistical Verdict**: Moderate positive boost associated with suburban growing environments.

---

### 5.2 Overall Model Performance

- **College Degree Accuracy (72.3%)**: The multinomial model correctly predicts college completion (Bachelor's or higher vs. non-college) for **72.3% of individuals**.
- **Continuous Schooling Accuracy (60.8%)**: The OLS model predicts total completed schooling within $\pm 2$ years for **60.8% of individuals** (Mean Absolute Error = 1.95 years).
- **Core Insight**: Parental education, family structure, sibling count, parental nativity, childhood income, and suburban geography form a robust, highly predictive framework for educational attainment in the United States.
