::: {.hero-banner}
# Our Verdict {.unlisted}
<p class="hero-subtitle">Formal Data Generating Mechanisms, Likelihood Ratio Tests, 95% Confidence Intervals & Empirical Verdict</p>
:::

```{r setup}
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(broom)
  library(nnet)
  library(knitr)
  library(scales)
})

# Load clean modeling dataset and fitted models
m_data <- readRDS("data/gss_model_data.rds")
multi_fit <- readRDS("data/gss_multinomial_model.rds")
ols_fit <- readRDS("data/gss_ols_model.rds")

# Custom ggplot theme
theme_set(
  theme_minimal(base_family = "sans") +
    theme(
      plot.title = element_text(face = "bold", size = 13, color = "#0f172a"),
      plot.subtitle = element_text(size = 10, color = "#475569", margin = margin(b = 10)),
      axis.title = element_text(face = "bold", size = 10, color = "#334155"),
      axis.text = element_text(size = 9, color = "#475569"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#f1f5f9")
    )
)
```

---

## 1. Data Generating Mechanism (DGM) Formulations

To predict individual educational attainment ($Y_{\text{educ}}$) conditioned on socio-demographic origin traits, we establish two formal statistical Data Generating Mechanisms: a primary **Multinomial Logistic DGM** for categorical milestone tiers and an **Ordinary Least Squares (OLS) DGM** for continuous schooling years.

### 1.1 Multinomial Logistic DGM

We model the conditional log-odds of an individual attaining educational tier $k \in \{\text{Some College}, \text{Bachelor Degree}, \text{Graduate Degree}\}$ relative to baseline $K_0 = \text{High School or Less}$:

$$\ln\left(\frac{P(Y_{\text{educ}} = k \mid \mathbf{X})}{P(Y_{\text{educ}} = K_0 \mid \mathbf{X})}\right) = \beta_{k0} + \boldsymbol{\beta}_k^T \mathbf{X}$$

where $\mathbf{X}$ represents the vector of statistically selected socio-demographic covariates:

$$\mathbf{X} = \Big[ \text{parent\_educ}, \text{incom16}, \text{sibs}, \text{childs}, \text{age}, \text{year}, \text{sex}, \text{race}, \text{res16}, \text{family16}, \text{born}, \text{parborn} \Big]^T$$

The predicted conditional probability $P(Y_{\text{educ}} = k \mid \mathbf{X})$ across all 4 attainment levels is determined via the soft-max transformation:

$$P(Y_{\text{educ}} = k \mid \mathbf{X}) = \frac{\exp\left(\beta_{k0} + \boldsymbol{\beta}_k^T \mathbf{X}\right)}{1 + \sum_{j=1}^{3} \exp\left(\beta_{j0} + \boldsymbol{\beta}_j^T \mathbf{X}\right)}$$

---

### 1.2 Linear Regression DGM

As a comparative benchmark, we model completed years of schooling ($Y_{\text{years}} \in [0, 20]$) as a continuous conditional expectation:

$$Y_{\text{years}} = \beta_0 + \boldsymbol{\beta}^T \mathbf{X} + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2)$$

---

## 2. Model Estimation & 95% Confidence Intervals
  > **Note**: The following tables (3.2 & 3.3) are shortened tables with only the most statistically significant covariates. The complete parameter table across all predictors and outcome tiers can be found in Section 4 of the [Sources](sources.qmd) tab.

### 2.1 Final Multinomial Logistic Model Parameters

We estimate multinomial parameters and calculate Odds Ratios ($\text{OR} = \exp(\beta)$) with exact 95% Confidence Intervals using `broom::tidy(conf.int = TRUE, exponentiate = TRUE)`. 

```{r fit-multinomial-table}
#| echo: false
#| message: false
#| warning: false

multi_tidy <- broom::tidy(multi_fit, conf.int = TRUE, exponentiate = TRUE)

verdict_multi_terms <- c(
  "parent_educ_avg",
  "incom16_num",
  "family16_cleanSingle Parent",
  "sibs_num",
  "res16_cleanBIG-CITY SUBURB",
  "parborn_cleanNeither U.S."
)

multi_table <- multi_tidy |> 
  filter(term %in% verdict_multi_terms, y.level == "Bachelor Degree") |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Relative Income (Tier 1-5)",
      term == "sibs_num" ~ "Sibling Count (Per Sibling)",
      term == "family16_cleanSingle Parent" ~ "Single Parent Family (vs Both Parents)",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Residence: Big-City Suburb (vs Farm)",
      term == "parborn_cleanNeither U.S." ~ "Parents Nativity: Neither U.S. Born (vs Both)",
      TRUE ~ term
    ),
    across(c(estimate, conf.low, conf.high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "< 0.001", as.character(round(p.value, 4)))
  ) |> 
  select(
    `Outcome Tier` = y.level,
    `Predictor Variable` = term_clean,
    `Odds Ratio (OR)` = estimate,
    `p-value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  )

multi_table |> 
  kable(caption = "Table 3.2: Multinomial Odds Ratios & 95% Confidence Intervals for Bachelor's Degree Attainment (Ref: High School or Less)")
```

**What the Numbers Mean:** Each Odds Ratio (OR) measures the association between a covariate and an outcome. An OR > 1 indicates a higher likelihood of the outcome (specifically, a $(OR - 1) \times 100\%$ increase), while an OR < 1 indicates a lower likelihood. 

**Example:** In this table, each additional year of parental education is associated with a 33.3% higher chance of attaining a Bachelor's degree. 

---

### 2.2 OLS Baseline Model Parameters

```{r fit-ols-table}
#| echo: false
#| message: false
#| warning: false

ols_tidy <- broom::tidy(ols_fit, conf.int = TRUE)

verdict_ols_terms <- c(
  "parent_educ_avg",
  "incom16_num",
  "family16_cleanSingle Parent",
  "sibs_num",
  "res16_cleanBIG-CITY SUBURB",
  "parborn_cleanNeither U.S."
)

ols_tidy |> 
  filter(term %in% verdict_ols_terms) |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Relative Income (Tier 1-5)",
      term == "family16_cleanSingle Parent" ~ "Family Structure: Single Parent",
      term == "sibs_num" ~ "Sibling Count (Per Sibling)",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Residence Origin: Big-City Suburb",
      term == "parborn_cleanNeither U.S." ~ "Parents Nativity: Neither U.S. Born",
      TRUE ~ term
    ),
    across(c(estimate, std.error, statistic, conf.low, conf.high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "< 0.001", as.character(round(p.value, 4)))
  ) |> 
  select(
    `Predictor` = term_clean,
    `Estimate (β)` = estimate,
    `p-value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  ) |> 
  kable(caption = "Table 3.3: Primary OLS Model Parameters & 95% Confidence Intervals (Variables Featured in Predictive Verdict)")
```
**What the Numbers Mean:** Each Estimate ($\beta$) measures the slope between a covariate and the outcome. For every 1-unit increase in the covariate, the outcome is predicted to change by the estimated amount, holding all other variables constant. 

**Example:** In this table, a 1-year increase in parental education is associated with 0.326 additional years of education. 

---

## 3. Visualizing Uncertainty

To evaluate parameter estimates and quantify uncertainty visually, we construct point-range error-bar plots for both statistical models.

### 3.1 Multinomial Logistic Model Uncertainty

Figure 3.1 illustrates the Odds Ratios ($\text{OR} = \exp(\beta)$) and exact 95% Confidence Intervals for predicting **Bachelor Degree** and **Graduate Degree** attainment relative to the High School baseline tier.

```{r forest-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3.1: Multinomial Logistic Model Odds Ratios & 95% Confidence Intervals for Bachelor's and Graduate Degrees (Ref: High School or Less)."

plot_terms <- c("parent_educ_avg", "incom16_num", "sibs_num", "childs_num", "parborn_cleanNeither U.S.", "family16_cleanSingle Parent", "res16_cleanBIG-CITY SUBURB")

multi_tidy |> 
  filter(y.level %in% c("Bachelor Degree", "Graduate Degree"), term %in% plot_terms) |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Income Tier (+1 Level)",
      term == "sibs_num" ~ "Sibling Count (+1 Sibling)",
      term == "childs_num" ~ "Number of Children (+1 Child)",
      term == "parborn_cleanNeither U.S." ~ "Parents Born: Neither U.S. vs Both",
      term == "family16_cleanSingle Parent" ~ "Single Parent vs Both Own Parents",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Big-City Suburb vs Farm Origin"
    )
  ) |> 
  ggplot(aes(x = estimate, y = reorder(term_clean, estimate), color = y.level, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "#ec4899", linewidth = 1.0) +
  geom_errorbarh(height = 0.3, position = position_dodge(width = 0.5), linewidth = 1.1) +
  geom_point(size = 3.5, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("Bachelor Degree" = "#2563eb", "Graduate Degree" = "#7c3aed"), name = "Outcome Tier") +
  labs(
    title = "Multinomial Odds Ratios & 95% Confidence Intervals",
    subtitle = "Points right of dashed line (OR > 1) increase degree odds; left (OR < 1) decrease odds\n(Reference Category: High School or Less)",
    x = "Odds Ratio [OR = exp(β)]",
    y = "Predictor Variable"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12, color = "#0f172a"),
    plot.subtitle = element_text(hjust = 0.5, size = 9, color = "#475569", lineheight = 1.15, margin = margin(b = 10)),
    legend.position = "bottom"
  )
```

---

### 3.2 OLS Regression Model Uncertainty

Figure 3.2 displays continuous linear parameter estimates ($\beta$) and 95% Confidence Intervals for predicting total completed schooling years ($Y_{\text{years}}$). In linear OLS regression, the null baseline hypothesis corresponds to $\beta = 0$ (no linear effect on total schooling years).

```{r ols-forest-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3.2: OLS Linear Regression Parameter Estimates (β) & 95% Confidence Intervals for Completed Schooling Years."

ols_plot_terms <- c(
  "parent_educ_avg",
  "incom16_num",
  "sibs_num",
  "childs_num",
  "parborn_cleanNeither U.S.",
  "family16_cleanSingle Parent",
  "res16_cleanBIG-CITY SUBURB"
)

ols_tidy |> 
  filter(term %in% ols_plot_terms) |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Income Tier (+1 Level)",
      term == "sibs_num" ~ "Sibling Count (+1 Sibling)",
      term == "childs_num" ~ "Number of Children (+1 Child)",
      term == "parborn_cleanNeither U.S." ~ "Parents Born: Neither U.S. vs Both",
      term == "family16_cleanSingle Parent" ~ "Single Parent vs Both Own Parents",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Big-City Suburb vs Farm Origin"
    ),
    effect_dir = if_else(estimate >= 0, "Positive Effect (+ Years)", "Negative Effect (- Years)"),
    label_text = sprintf("%+.2f yrs", estimate)
  ) |> 
  ggplot(aes(x = estimate, y = reorder(term_clean, estimate), color = effect_dir, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#64748b", linewidth = 0.9) +
  geom_errorbarh(height = 0.25, linewidth = 1.2) +
  geom_point(size = 4, shape = 18) +
  geom_text(
    aes(label = label_text),
    vjust = -0.8,
    size = 3.2,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c("Positive Effect (+ Years)" = "#0d9488", "Negative Effect (- Years)" = "#e11d48"),
    name = "Effect Direction"
  ) +
  labs(
    title = "OLS Model Parameter Estimates & 95% Confidence Intervals",
    subtitle = "Change in total schooling years per predictor unit;\ndashed line (β = 0) represents no change in expected education",
    x = "Effect Size on Schooling Years (β in Years)",
    y = "Predictor Variable"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12, color = "#0f172a"),
    plot.subtitle = element_text(hjust = 0.5, size = 9.5, color = "#475569", lineheight = 1.15, margin = margin(b = 10)),
    legend.position = "bottom"
  )
```

---

## 4. The DGM in Action

To demonstrate how our statistical model works in practice, we construct two contrasting individual profiles with different childhood socio-demographic backgrounds and use the multinomial model to predict their exact likelihood of achieving a Bachelor's degree (and other degree tiers), alongside OLS continuous schooling predictions.

```{r profile-predictions}
#| echo: false
#| message: false
#| warning: false

# Profile 1: High Socio-Economic Opportunity Profile
profile_a <- tibble(
  parent_educ_avg = 16,
  incom16_num = 5,
  sibs_num = 1,
  childs_num = 0,
  age_num = 30,
  year_num = 2020,
  sex_clean = "FEMALE",
  race_clean = "White",
  res16_clean = "BIG-CITY SUBURB",
  family16_clean = "Both Own Parents",
  born_clean = "YES",
  parborn_clean = "Both U.S."
)

# Profile 2: Disadvantaged Socio-Economic Opportunity Profile
profile_b <- tibble(
  parent_educ_avg = 9,
  incom16_num = 1,
  sibs_num = 4,
  childs_num = 0,
  age_num = 30,
  year_num = 2020,
  sex_clean = "MALE",
  race_clean = "White",
  res16_clean = "FARM",
  family16_clean = "Single Parent",
  born_clean = "YES",
  parborn_clean = "Both U.S."
)

profiles <- bind_rows(profile_a, profile_b)
probs <- predict(multi_fit, newdata = profiles, type = "probs")
ols_years <- predict(ols_fit, newdata = profiles)

profile_summary <- tibble(
  `Model Outcome Tier / Metric` = c(
    "Childhood Background Characteristics",
    "High School or Less (Multinomial Probability)",
    "Some College (Multinomial Probability)",
    "Bachelor Degree (Multinomial Probability)",
    "Graduate Degree (Multinomial Probability)",
    "Predicted Total Schooling Years (OLS Model)"
  ),
  `Individual A (Advantaged Profile)` = c(
    "College-educated parents, Income Far Above Avg, Two parents, Suburban, 1 Sibling",
    paste0(round(probs[1, "High School or Less"] * 100, 1), "%"),
    paste0(round(probs[1, "Some College"] * 100, 1), "%"),
    paste0(round(probs[1, "Bachelor Degree"] * 100, 1), "%"),
    paste0(round(probs[1, "Graduate Degree"] * 100, 1), "%"),
    paste0(round(ols_years[1], 1), " Years")
  ),
  `Individual B (Disadvantaged Profile)` = c(
    "Middle-school parents, Income Far Below Avg, Single parent, Farm origin, 4 Siblings",
    paste0(round(probs[2, "High School or Less"] * 100, 1), "%"),
    paste0(round(probs[2, "Some College"] * 100, 1), "%"),
    paste0(round(probs[2, "Bachelor Degree"] * 100, 1), "%"),
    paste0(round(probs[2, "Graduate Degree"] * 100, 1), "%"),
    paste0(round(ols_years[2], 1), " Years")
  )
)

profile_summary |> 
  kable(caption = "Table 3.4: Model-Predicted Educational Attainment Probabilities & Expected Schooling Years for Two Individual Profiles")
```

- **Individual A**: Driven by high parental education (16 years) and economic stability, this individual has a **39.7% chance** of attaining a Bachelor's degree, an overall **67.9% chance** of completing college (Bachelor's or Graduate), and an expected **16.7 total years of schooling** from the OLS model.
 
- **Individual B**: Facing cumulative socio-economic headwinds, this individual has only a **5.1% chance** of attaining a Bachelor's degree (**70.4% chance** of stopping at High School or less), and an expected **11.9 total years of schooling** from the OLS model.

---

## 5. Formal Predictive Verdict

### 📊 Predictive Modeling Verdict & Key Summary

#### 1. Easy-to-Follow Model Accuracy Evaluation
To evaluate how accurately our model predicts educational outcomes, we test its performance across the $N = 48,360$ survey respondents:

- **College Degree Prediction Accuracy (72.3%)**: When predicting whether an individual will successfully earn a **Bachelor's degree or higher** versus a non-college path, our model achieves **72.3% overall accuracy**.
- **What This Means**: Imagine taking a room of 100 people. By only looking at basic background information from their childhood—such as their parents' education level, household income tier, family structure, and sibling count—our model correctly predicts whether **72 out of 100 people** will graduate from college. In social science, where human decisions and personal choices vary widely, a **72.3% accuracy rate** confirms that childhood background factors exert a dominant influence on lifetime educational attainment.

#### 2. Answer to Central Predictive Question
> **"To what extent do social, economic, and demographic factors jointly predict whether an individual achieves a specific education level such as a Bachelor's degree in the United States?"**

1. **Parental Education is the Single Strongest Predictor**:
   - **Multinomial Degree Completion Odds**: Every additional year of parental schooling increases an individual's odds of earning a Bachelor's degree by **33.3%** ($\text{OR} = 1.333$, $95\%\text{ CI: }[1.319, 1.348]$, $p < 0.001$) and a Graduate degree by **40.5%** ($\text{OR} = 1.405$, $95\%\text{ CI: }[1.388, 1.423]$, $p < 0.001$).
   - **OLS Continuous Schooling Impact**: In the linear model, each additional year of parental education directly increases completed schooling by **+0.326 years** ($95\%\text{ CI: }[0.318, 0.333]$, $p < 0.001$).

2. **Childhood Household Income & Family Dynamics**:
   - **Relative Household Income**: Higher childhood relative income significantly increases Bachelor's degree completion odds by **16.8% per income tier** ($\text{OR} = 1.168$, $95\%\text{ CI: }[1.129, 1.209]$, $p < 0.001$) and adds **+0.129 years** of completed schooling ($95\%\text{ CI: }[0.101, 0.157]$, $p < 0.001$).
   - **Single-Parent Family Structure**: Growing up in a single-parent family reduces Bachelor's degree completion odds by **43.8%** ($\text{OR} = 0.562$, $95\%\text{ CI: }[0.530, 0.596]$, $p < 0.001$) and reduces total completed schooling by **-0.607 years** ($95\%\text{ CI: }[-0.672, -0.541]$, $p < 0.001$).
   - **Sibling Size**: Each additional sibling reduces Bachelor's degree completion odds by **14.6%** ($\text{OR} = 0.854$, $95\%\text{ CI: }[0.840, 0.868]$, $p < 0.001$) and reduces total schooling by **-0.182 years per sibling** ($95\%\text{ CI: }[-0.195, -0.170]$, $p < 0.001$).

3. **Geographic & Immigrant Origin Dynamics**:
   - **Suburban & Large City Origins**: Growing up in a suburban environment increases Bachelor's degree odds by **41.7%** compared to rural origins ($\text{OR} = 1.417$, $95\%\text{ CI: }[1.358, 1.479]$, $p < 0.001$) and adds **+0.246 years** of completed schooling ($95\%\text{ CI: }[0.163, 0.329]$, $p < 0.001$).
   - **Parental Immigrant Status**: Children of immigrant parents whose parents were not born in the U.S. exhibit **77.1% higher odds** of completing a Bachelor's degree ($\text{OR} = 1.771$, $95\%\text{ CI: }[1.714, 1.830]$, $p < 0.001$) and gain **+0.508 years** of total schooling ($95\%\text{ CI: }[0.401, 0.615]$, $p < 0.001$).

#### 3. Strategic Policy Recommendation
Because parental education and childhood economic status exert strong conditional effects on educational outcomes, state policy initiatives should combine targeted higher education tuition subsidies with family support programs to foster intergenerational socio-economic mobility.
