---
title: "Data Sources & References"
format:
  html:
    toc: true
    toc-depth: 2
---

::: {.hero-banner}
# Data Sources & References {.unlisted}
<p class="hero-subtitle">Data Provenance, Variable Specifications, and Full Parameter Tables</p>
:::

## 1. Primary Data Source

The empirical analysis in this project uses data from the **General Social Survey (GSS)**, conducted by **NORC at the University of Chicago**.

### General Social Survey (NORC at the University of Chicago)

- **APA Reference Citation**:
  > Davern, M., Bautista, R., Freese, J., Herd, P., & Morgan, S. L. (2024). *General Social Survey 1972–2024 (Release 3)* [Data set]. NORC at the University of Chicago. [https://gssdataexplorer.norc.org](https://gssdataexplorer.norc.org)

- **Methodology & Scope**:
  The General Social Survey (GSS) monitors societal trends and demographic transformations across the adult, non-institutionalized U.S. population using nationally representative probability sampling.

---

## 2. Variable Codebook & Definitions

Below are the official GSS codebook definitions for all 15 variables analyzed in this study:

| Variable Name | GSS Label | Description & Coding Scheme |
| :--- | :--- | :--- |
| **`educ`** | Respondent Educational Attainment | Highest grade completed (0–20 years). Recoded into: *High School or Less*, *Some College*, *Bachelor Degree*, *Graduate Degree*. |
| **`paeduc`** | Father's Educational Attainment | Highest grade completed by father (0–20 years). |
| **`maeduc`** | Mother's Educational Attainment | Highest grade completed by mother (0–20 years). |
| **`incom16`** | Relative Family Income at Age 16 | Income at age 16 relative to U.S. families: 1 = Far Below Average to 5 = Far Above Average. |
| **`family16`** | Family Structure at Age 16 | Living arrangement at age 16: *Both Own Parents*, *Single Parent*, *Stepparent*, *Other Relative*. |
| **`res16`** | Residence Type at Age 16 | Type of place lived in at age 16: *Farm*, *Country Non-Farm*, *Town <50k*, *50k to 250k*, *Big-City Suburb*, *City >250k*. |
| **`sex`** | Respondent's Sex | Sex assigned at interview: *Female*, *Male*. |
| **`race`** | Respondent's Race | Race of respondent: *White*, *Black*, *Other*. |
| **`born`** | Respondent Born in U.S. | Born in the United States: *Yes*, *No*. |
| **`parborn`** | Parents Born in U.S. | Parents born in U.S.: *Both U.S.*, *One U.S.*, *Neither U.S.* |
| **`sibs`** | Number of Siblings | Total number of brothers and sisters (0 to 6+). |
| **`childs`** | Number of Children | Total number of children (0 to 8+). |
| **`age`** | Respondent's Age | Age at interview (18 to 89 years). |
| **`year`** | Survey Administration Year | Survey administration year (1972 to 2024). |
| **`agekdbrn`** | Age at First Child's Birth | Age when first child was born. Excluded from final model due to missingness across early survey rounds. |

---

## 3. Appendix: Complete Model Parameter Tables

Below are the complete parameter estimation tables for both fitted statistical models.

```{r setup-appendix}
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(dplyr)
  library(nnet)
  library(broom)
  library(knitr)
})

m_data <- readRDS("data/gss_model_data.rds")
multi_fit <- readRDS("data/gss_multinomial_model.rds")
ols_fit <- readRDS("data/gss_ols_model.rds")

multi_tidy <- broom::tidy(multi_fit, conf.int = TRUE, exponentiate = TRUE)
ols_tidy <- broom::tidy(ols_fit, conf.int = TRUE)
```

### 3.1 Full Multinomial Logistic Model Parameters & Odds Ratios

Complete table of estimated parameters, Odds Ratios ($\text{OR} = \exp(\beta)$), and 95% Confidence Intervals relative to baseline *High School or Less*.

```{r fit-multinomial-full-table}
#| echo: false
#| message: false
#| warning: false

multi_tidy |> 
  filter(term != "(Intercept)") |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Relative Income (Tier 1-5)",
      term == "sibs_num" ~ "Sibling Count",
      term == "childs_num" ~ "Number of Children",
      term == "age_num" ~ "Respondent Age (Years)",
      term == "year_num" ~ "Survey Administration Year",
      term == "sex_cleanMALE" ~ "Sex: Male (vs Female)",
      term == "race_cleanWhite" ~ "Race: White (vs Black)",
      term == "race_cleanOther" ~ "Race: Other (vs Black)",
      term == "family16_cleanSingle Parent" ~ "Family Structure: Single Parent",
      term == "family16_cleanStepparent" ~ "Family Structure: Stepparent",
      term == "family16_cleanOther Relative" ~ "Family Structure: Other Relative",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Residence: Big-City Suburb",
      term == "res16_cleanCITY GT 250000" ~ "Residence: Large City >250k",
      term == "res16_cleanCOUNTRY,NONFARM" ~ "Residence: Rural Non-Farm",
      term == "res16_cleanTOWN LT 50000" ~ "Residence: Town <50k",
      term == "res16_cleanFARM" ~ "Residence: Farm",
      term == "born_cleanYES" ~ "Respondent U.S. Born",
      term == "parborn_cleanNeither U.S." ~ "Parents Nativity: Neither U.S.",
      term == "parborn_cleanOne U.S." ~ "Parents Nativity: One U.S.",
      TRUE ~ term
    ),
    across(c(estimate, conf.low, conf.high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "< 0.001", as.character(round(p.value, 4)))
  ) |> 
  select(
    `Outcome Tier` = y.level,
    `Predictor Variable` = term_clean,
    `Odds Ratio (OR)` = estimate,
    `p-value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  ) |> 
  kable(caption = "Table 3.1: Full Multinomial Logistic Model Parameters & Odds Ratios across All Outcome Tiers")
```

---

### 3.2 Full OLS Baseline Model Parameters

Complete table of estimated parameters from the OLS linear regression model for completed schooling years ($Y_{\text{years}}$).

```{r fit-ols-full-table}
#| echo: false
#| message: false
#| warning: false

ols_tidy |> 
  filter(term != "(Intercept)") |> 
  mutate(
    term_clean = case_when(
      term == "parent_educ_avg" ~ "Parental Education (Per Year)",
      term == "incom16_num" ~ "Childhood Relative Income (Tier 1-5)",
      term == "sibs_num" ~ "Sibling Count",
      term == "childs_num" ~ "Number of Children",
      term == "age_num" ~ "Respondent Age (Years)",
      term == "year_num" ~ "Survey Administration Year",
      term == "sex_cleanMALE" ~ "Sex: Male",
      term == "race_cleanWhite" ~ "Race: White",
      term == "race_cleanOther" ~ "Race: Other",
      term == "family16_cleanSingle Parent" ~ "Family Structure: Single Parent",
      term == "family16_cleanStepparent" ~ "Family Structure: Stepparent",
      term == "family16_cleanOther Relative" ~ "Family Structure: Other Relative",
      term == "res16_cleanBIG-CITY SUBURB" ~ "Residence Origin: Big-City Suburb",
      term == "res16_cleanCITY GT 250000" ~ "Residence Origin: Large City >250k",
      term == "res16_cleanCOUNTRY,NONFARM" ~ "Residence Origin: Rural Non-Farm",
      term == "res16_cleanTOWN LT 50000" ~ "Residence Origin: Town <50k",
      term == "res16_cleanFARM" ~ "Residence Origin: Farm",
      term == "born_cleanYES" ~ "Respondent U.S. Born",
      term == "parborn_cleanNeither U.S." ~ "Parents Nativity: Neither U.S.",
      term == "parborn_cleanOne U.S." ~ "Parents Nativity: One U.S.",
      TRUE ~ term
    ),
    across(c(estimate, std.error, statistic, conf.low, conf.high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "< 0.001", as.character(round(p.value, 4)))
  ) |> 
  select(
    `Predictor` = term_clean,
    `Estimate (β)` = estimate,
    `Std. Error` = std.error,
    `t statistic` = statistic,
    `p-value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  ) |> 
  kable(caption = "Table 3.2: Full OLS Model Parameters & 95% Confidence Intervals for Completed Schooling Years")
```
